---
title: "SOEP Data Preparation"
author: "Salma-Anastasiia-Suzanna"
date: "2024-12-04"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    df_print: paged
---
# Introduction
We are examining the extent to which individuals’ subjective well-being can be predicted by objective factors that determine individual health decline and for which group this declaration bias is particularly large. And in that sense, which predictors (such as age, labor market status, personality traits and hand-grip strength) help improve predictions of individual health declines.

This report documents the steps taken to preprocess and analyze the SOEP data for health-related outcomes: subjective and objective health measures, predictors, and control variables. The transformations include cleaning, imputing missing values, and deriving new variables.

------------------------------------------------------------------------

# 1. Setup

## 1.1 Load Required Packages

```{r}
# Load required packages
my_packages <- c("tidyverse", "haven", "openxlsx", "labelled", "dplyr", "lubridate","ggplot2","skimr")

for (p in my_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, dependencies = TRUE)
  }
  library(p, character.only = TRUE)
}
```

## 1.2 Load the pl Dataset

```{r}
main <- read_dta('/Users/salmahoumane/Documents/R/pl.dta') %>%
  select(
    "pid", "syear", "ple0008", "plh0035", "ple0055", "ple0056", "ple0072", "ple0081_h", "ple0086_v4",
    "ple0080_v2", "ple0086_v1", "ple0086_v4", paste0("ple00", 11:23),
    "ple0004", "ple0005", "ple0095", "plh0182", "plh0171",
    "plh0042", "plh0033", "plh0032", "ple0010_h", "ple0006", "ple0007"
  )
cat("Main dataset dimensions:", dim(main), "\n")
```

## 1.3 Merge with Other Datasets

```{r}
# Merge with `pgen` dataset
data_pgen <- read_dta('/Users/salmahoumane/Documents/R/pgen.dta') %>%
  select("pid", "syear", "pgerwzeit", "pgemplst", "pglabnet", "pgtatzeit", "pgexpue", "pgstib","pgfamstd", "pgisced11")
cat("PGEN dataset dimensions:", dim(data_pgen), "\n")
main_before_merge <- dim(main)  # Store dimensions before merge
main <- merge(main, data_pgen, by = c("pid", "syear"), all.x = TRUE)
cat("After merging with pgen: Rows =", nrow(main), "Cols =", ncol(main), "\n")
cat("Same number of rows then the main dataset with 7 new rows -> Merge successful")

# Check for duplicate rows introduced during merge
duplicates <- main %>%
  group_by(pid, syear) %>%
  summarise(count = n()) %>%
  filter(count > 1)

if (nrow(duplicates) > 0) {
  cat("Warning: Duplicates detected after merging with pgen!\n")
  print(duplicates)
} else {
  cat("No duplicates found after merging with pgen.\n")
}

# Merge with `ppathl` dataset
data_ppathl <- read_dta('/Users/salmahoumane/Documents/R/ppathl.dta') %>%
  select("pid", "hid", "syear", "psample", "gebjahr", "sex", "migback", "germborn")
cat("PPATHL dataset dimensions:", dim(data_ppathl), "\n")
main_before_ppathl <- dim(main)  # Store dimensions before second merge

main <- left_join(main, data_ppathl, by = c("pid", "syear"))
cat("After merging with ppathl: Rows =", nrow(main), "Cols =", ncol(main), "\n")
cat("Same number of rows then the main dataset with 6 new rows -> Merge successful")

# Final snapshot of dimensions
cat("Final dataset dimensions after all merges:", dim(main), "\n")
```

## 1.4 Main filtering: Analysis post 1999

```{r}
# Drop observations prior to 1999
main <- main %>%
  filter(syear >= 1999)
cat("Dimensions:", dim(main), "\n")
```

# 2. Outcome Variables: Subjective Health

## 2.1 Health

```{r}
# Clean and rename the `health` variable
main <- main %>%
  rename(health = ple0008) %>%
  filter(health != -1) %>%  # Filter valid health values -> 398 observations dropped
    mutate(health = case_when(health == 5 ~ 1, # Reverse the scale: 1 -> 5, 5 -> 1
                            health == 4 ~ 2,
                            health == 3 ~3,
                            health == 2 ~ 4,
                            health == 1 ~ 5))

sum(is.na(main$health))  # no missings

# Visualisation: Bar plot
ggplot(main, aes(x = factor(health))) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Self-Assessed Health",
    x = "Health Category (1 = Poor, 5 = Excellent)",
    y = "Count"
  ) +
  theme_minimal()
```

## 2.2 Worries About Health

```{r}
# Create `worr_health_categorical` and `worr_health_dummy`
main <- main %>%
  rename(worr_health_categorical = plh0035) %>%
  filter(worr_health_categorical != -1 & worr_health_categorical!= -4) %>% # drop those who refused to answer -> 1287 obs
  mutate(
    worr_health_categorical = case_when(
      worr_health_categorical < 0 ~ NA_real_,
      TRUE ~ worr_health_categorical
    )
  ) %>%
  group_by(pid) %>%
  mutate(
    worr_health_categorical = if_else(
      syear %in% c(2011, 2012, 2013) & is.na(worr_health_categorical), #impute the years 2011-2013 because 11000 observations werent asked this question in these years
      floor(
        mean(
          worr_health_categorical[syear %in% c(2010, 2014)],
          na.rm = TRUE
        )
      ),
      worr_health_categorical
    )
  ) %>%
  ungroup() %>%
  drop_na(worr_health_categorical) %>% # 1032
  mutate(worr_health_categorical = case_when(
                            worr_health_categorical == 3 ~ 1, # Reverse the scale: 1 = no worries
                            worr_health_categorical == 2 ~ 2,
                            worr_health_categorical == 1 ~ 3), # 3 = lots of worries
    worr_health_dummy = if_else(
      worr_health_categorical >= 2, 1, 0
    )
  )

# Visualisation: Bar plot (categorical)
ggplot(main, aes(x = worr_health_categorical)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Distribution of Worries About Health",
    x = "Worries About Health",
    y = "Count"
  ) +
  theme_minimal()
# Visualisation: Pie Char (dummy)
main %>%
  group_by(worr_health_dummy) %>%
  summarize(count = n()) %>%
  mutate(
    proportion = count / sum(count),
    label = paste0(round(proportion * 100, 1), "%")
  ) %>%
  ggplot(aes(x = "", y = proportion, fill = factor(worr_health_dummy))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(
    title = "Proportion of Health Worry (Dummy)",
    fill = "Health Worry (Dummy)"
  ) +
  theme_minimal() +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5))
```

## 2.3 Health Trends

```{r}
# Create health decline dummy for 2 years into the future
main <- main %>%
  arrange(pid, syear) %>%
  group_by(pid) %>%
  mutate(
    health_in_2yrs = lead(health, n = 2, order_by = syear),
    health_decline_2yrs = case_when(
      !is.na(health_in_2yrs) & health_in_2yrs < health ~ 1,
      !is.na(health_in_2yrs) & health_in_2yrs >= health ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()
sum(is.na(main$health_in_2yrs)) # 81.113
#Visualisation: Line plot
# Line plot: Average health over years
main %>%
  group_by(syear) %>%
  summarize(avg_health = mean(health, na.rm = TRUE)) %>%
  ggplot(aes(x = syear, y = avg_health)) +
  geom_line(color = "darkgreen") +
  geom_point(color = "darkgreen") +
  labs(
    title = "Average Self-Assessed Health Over Years",
    x = "Survey Year",
    y = "Average Health"
  ) +
  theme_minimal()
```

# 3. Predictors
## 3.2 Life Satisfaction and Health Satisfaction
```{r}
# Create life_satisfaction and health_satisfaction variables
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    plh0182_NA = sum(is.na(plh0182)),
    plh0182_not_asked = sum(plh0182 %in% c(-8, -5), na.rm = TRUE),
    plh0182_other_neg = sum(plh0182 < 0 & !(plh0182 %in% c(-8, -5)), na.rm = TRUE),
    plh0171_NA = sum(is.na(plh0171)),
    plh0171_not_asked = sum(plh0171 %in% c(-8, -5), na.rm = TRUE),
    plh0171_other_neg = sum(plh0171 < 0 & !(plh0171 %in% c(-8, -5)), na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Drop those who refused to answer or answered invalidly
main <- main %>%
  filter(!(plh0182 %in% c(-1, -2, -3) | plh0171 %in% c(-1, -2, -3))) # 908 observations

# Step 3: Impute remaining Values for plh0182 and plh0171 with Max 2-Year Lag/Lead
main <- main %>%
  arrange(pid, syear) %>%  # Ensure proper sorting
  group_by(pid) %>%
  mutate(
    # Flag rows that will be imputed
    imputed_flag_life = if_else(plh0182 %in% c(-5, -8) | is.na(plh0182), 1, 0),
    imputed_flag_health = if_else(plh0171 %in% c(-5, -8) | is.na(plh0171), 1, 0),

    # Replace -5 and -8 with NA
    plh0182 = case_when(plh0182 %in% c(-5, -8) ~ NA_real_, TRUE ~ plh0182),
    plh0171 = case_when(plh0171 %in% c(-5, -8) ~ NA_real_, TRUE ~ plh0171),

    # Impute plh0182 (life satisfaction)
    life_satisfaction = if_else(
      is.na(plh0182),
      coalesce(
        if_else(syear - lag(syear) <= 2 & !is.na(lag(plh0182)), lag(plh0182), NA_real_),
        if_else(lead(syear) - syear <= 2 & !is.na(lead(plh0182)), lead(plh0182), NA_real_)
      ),
      plh0182
    ),

    # Impute plh0171 (health satisfaction)
    health_satisfaction = if_else(
      is.na(plh0171),
      coalesce(
        if_else(syear - lag(syear) <= 2 & !is.na(lag(plh0171)), lag(plh0171), NA_real_),
        if_else(lead(syear) - syear <= 2 & !is.na(lead(plh0171)), lead(plh0171), NA_real_)
      ),
      plh0171
    )
  ) %>%
  ungroup()

# Step 3: Breakdown After Imputation
missing_breakdown_after <- main %>%
  summarise(
    life_satisfaction_NA = sum(is.na(life_satisfaction)),
    health_satisfaction_NA = sum(is.na(health_satisfaction))
  )

print("Breakdown of missing values after imputation:")
print(missing_breakdown_after)

# Drop observations that are still missing
main <- main %>%
  filter(!is.na(life_satisfaction) & !is.na(health_satisfaction)) # 681 observations dropped

# Step 4: Descriptive Statistics
# Imputed Rows
imputed_stats <- main %>%
  filter(imputed_flag_life == 1 | imputed_flag_health == 1) %>%
  summarise(
    Life_Satisfaction_Mean = mean(life_satisfaction, na.rm = TRUE),
    Life_Satisfaction_Median = median(life_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Mean = mean(health_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Median = median(health_satisfaction, na.rm = TRUE),
    Count = n()
  )

# Non-Imputed Rows
non_imputed_stats <- main %>%
  filter(imputed_flag_life == 0 & imputed_flag_health == 0) %>%
  summarise(
    Life_Satisfaction_Mean = mean(life_satisfaction, na.rm = TRUE),
    Life_Satisfaction_Median = median(life_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Mean = mean(health_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Median = median(health_satisfaction, na.rm = TRUE),
    Count = n()
  )

print("Descriptive Statistics for Imputed Rows:")
print(imputed_stats)

print("Descriptive Statistics for Non-Imputed Rows:")
print(non_imputed_stats)

# The imputed observations display slighly larger health & life-satisfaction

# Step 5: Visualizations
# All Rows: Life Satisfaction
ggplot(main, aes(x = factor(life_satisfaction))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Life Satisfaction - All Rows",
    x = "Life Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Imputed Rows Only
ggplot(main %>% filter(imputed_flag_life == 1), aes(x = factor(life_satisfaction))) +
  geom_bar(fill = "orange", color = "black") +
  labs(
    title = "Life Satisfaction - Imputed Rows",
    x = "Life Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Non-Imputed Rows Only
ggplot(main %>% filter(imputed_flag_life == 0), aes(x = factor(life_satisfaction))) +
  geom_bar(fill = "green", color = "black") +
  labs(
    title = "Life Satisfaction - Non-Imputed Rows",
    x = "Life Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# All Rows: Health Satisfaction
ggplot(main, aes(x = factor(health_satisfaction))) +
  geom_bar(fill = "coral", color = "black") +
  labs(
    title = "Health Satisfaction - All Rows",
    x = "Health Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Imputed Rows Only
ggplot(main %>% filter(imputed_flag_health == 1), aes(x = factor(health_satisfaction))) +
  geom_bar(fill = "orange", color = "black") +
  labs(
    title = "Health Satisfaction - Imputed Rows",
    x = "Health Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Non-Imputed Rows Only
ggplot(main %>% filter(imputed_flag_health == 0), aes(x = factor(health_satisfaction))) +
  geom_bar(fill = "green", color = "black") +
  labs(
    title = "Health Satisfaction - Non-Imputed Rows",
    x = "Health Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()
```

## 3.2.2 Height, Weight, BMI
```{r}
# Process height and weight
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    height_NA = sum(is.na(ple0006)),
    height_not_asked = sum(ple0006 %in% c(-8, -5), na.rm = TRUE),
    height_invalid = sum(ple0006 > 245 | (ple0006 < 0 & !(ple0006 %in% c(-8, -5))), na.rm = TRUE),
    weight_NA = sum(is.na(ple0007)),
    weight_not_asked = sum(ple0007 %in% c(-8, -5), na.rm = TRUE),
    weight_invalid = sum(ple0007 < 0 & !(ple0007 %in% c(-8, -5)), na.rm = TRUE)
  )

print("Breakdown of missing and invalid values before imputation:")
print(missing_breakdown_before)

# Step 2: Process and Impute Height with Lag/Lead (<= 2 years)
  # Assumption for Imputation of height: does not change significantly for adults
main <- main %>%
  arrange(pid, syear) %>%
  group_by(pid) %>%
  mutate(
    # Track imputation flag for height
    height_imputed = if_else(ple0006 > 245 | ple0006 <= 0 | is.na(ple0006), 1, 0),
    # Set invalid or missing height values to missing
    height = ifelse(ple0006 < 0 | ple0006 > 245, NA, ple0006),
    # Forward fill missing values
    height = zoo::na.locf(height, na.rm = FALSE),

    # Backward fill missing values
    height = zoo::na.locf(height, na.rm = FALSE, fromLast = TRUE)
  ) %>%
  ungroup()

# Step 3: Process and Impute Weight with Lag/Lead (<= 2 years)
#imputation: The mean weight between the last observation before the missing and the first after. And for the ones that have only one before or one after, the one but only if it is an adjacent year.
main <- main %>%
  filter(!(ple0007 %in% c(-1, -2, -3))) %>%
  group_by(pid) %>%
  arrange(syear, .by_group = TRUE) %>%
  mutate(# Track imputation flag for height
    weight_imputed = if_else(ple0007 <= 0 | is.na(ple0007), 1, 0),
    ple0007 = case_when(ple0007<0 ~ NA_real_,
                             TRUE ~ ple0007), 
    # Create lag and lead weights
    lag_weight = lag(ple0007),
    lead_weight = lead(ple0007),
    lag_year = lag(syear),
    lead_year = lead(syear),
    
    # Apply imputation rules
    weight = case_when(
      is.na(ple0007) & !is.na(lag_weight) & !is.na(lead_weight) & 
        (syear - lag_year == 1 & lead_year - syear == 1) ~ (lag_weight + lead_weight) / 2, # Mean of valid neighbors
      is.na(ple0007) & !is.na(lag_weight) & (syear - lag_year == 1) ~ lag_weight,         # Use previous value if adjacent
      is.na(ple0007) & !is.na(lead_weight) & (lead_year - syear == 1) ~ lead_weight,     # Use next value if adjacent
      TRUE ~ ple0007  # Keep original if no condition is met
    )
  ) %>%
  select(-lag_weight, -lead_weight, -lag_year, -lead_year) %>%  # Clean up temporary columns
  ungroup()



# Step 4: Track Missing Categories After Imputation
missing_breakdown_after <- main %>%
  summarise(
    height_NA = sum(is.na(height)),
    weight_NA = sum(is.na(weight)),
    height_imputed = sum(height_imputed),
    weight_imputed = sum(weight_imputed)
  )

print("Breakdown of missing values after imputation:")
print(missing_breakdown_after)

main <- main %>%
  filter(!is.na(height) & !is.na(weight))

# Step 5: Calculate BMI and Create Categorical Variable
main <- main %>%
  mutate(
    BMI = as.numeric(weight / ((height / 100) ^ 2)),  # Calculate BMI
    bmi_categorical = case_when(
      BMI < 18.5 ~ 1, #underweight
      BMI >= 18.5 & BMI < 25 ~ 2, #normal
      BMI >= 25 & BMI < 30 ~ 3, #overweight
      BMI >= 30 ~ 4, # obese
      TRUE ~ NA_real_
    )
  )

# Step 6: Descriptive Statistics for Imputed and Non-Imputed Rows
# Descriptive stats for all rows
all_stats <- main %>%
  summarise(
    Mean = mean(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE),
    Min = min(BMI, na.rm = TRUE),
    Max = max(BMI, na.rm = TRUE),
    Count = n()
  )

# Descriptive stats for imputed rows
imputed_stats <- main %>%
  filter(height_imputed == 1 | weight_imputed == 1) %>%
  summarise(
    Mean = mean(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE),
    Min = min(BMI, na.rm = TRUE),
    Max = max(BMI, na.rm = TRUE),
    Count = n()
  )

# Descriptive stats for non-imputed rows
non_imputed_stats <- main %>%
  filter(height_imputed == 0 & weight_imputed == 0) %>%
  summarise(
    Mean = mean(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE),
    Min = min(BMI, na.rm = TRUE),
    Max = max(BMI, na.rm = TRUE),
    Count = n()
  )

print("Descriptive Statistics for All Rows:")
print(all_stats)

print("Descriptive Statistics for Imputed Rows:")
print(imputed_stats)

print("Descriptive Statistics for Non-Imputed Rows:")
print(non_imputed_stats)

# Step 7: Visualizations
# Histogram for all rows
ggplot(main, aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black", alpha = 0.7) +
  labs(
    title = "BMI Distribution - All Rows",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()

# Histogram for imputed rows
ggplot(main %>% filter(height_imputed == 1 | weight_imputed == 1), aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "orange", color = "black", alpha = 0.7) +
  labs(
    title = "BMI Distribution - Imputed Rows",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()

# Histogram for non-imputed rows
ggplot(main %>% filter(height_imputed == 0 & weight_imputed == 0), aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "green", color = "black", alpha = 0.7) +
  labs(
    title = "BMI Distribution - Non-Imputed Rows",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()

```


## 3.2.3 Smoking
Number of Cigarettes has too little observations
```{r}
# Create `smoking_dummy`
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    ple0081_h_NA = sum(is.na(ple0081_h)),
    ple0081_h_not_asked = sum(ple0081_h %in% c(-5, -8), na.rm = TRUE),
    ple0081_h_invalid = sum(ple0081_h < 0 & !(ple0081_h %in% c(-8, -5)), na.rm = TRUE)
  )

print("Breakdown of missing and invalid values before imputation:")
print(missing_breakdown_before)

# Step 2: Drop those who refused to answer or answered invalidly
main <- main %>%
  filter(!(ple0081_h %in% c(-1, -3) | ple0086_v4 %in% c(-1, -3))) # 137 observations

# Step 3.1: Impute Smoking Status (ple0081_h) with Lag and Lead (Max 2 Years)
main <- main %>%
  arrange(pid, syear) %>%
  group_by(pid) %>%
  mutate(
    # Flag missing smoking values
    smoking_imputed = if_else(ple0081_h %in% c(-5, -8), 1, 0),

    # Determine if the individual was ever observed to smoke
    ever_smoked = any(ple0081_h == 1, na.rm = TRUE),

    # Impute smoking values
    ple0081_h = case_when(
      # If missing and the person ever smoked, impute using lag/lead within 2 years
      ple0081_h %in% c(-5, -8) & ever_smoked ~ {
        lag_val <- if_else(syear - lag(syear) <= 2, lag(ple0081_h), NA_real_)
        lead_val <- if_else(lead(syear) - syear <= 2, lead(ple0081_h), NA_real_)
        if_else(
          !is.na(lag_val) & !is.na(lead_val),
          (lag_val + lead_val) / 2,
          coalesce(lag_val, lead_val)
        )
      },
      # If missing and the person never smoked, set to 2
      ple0081_h %in% c(-5, -8) & !ever_smoked ~ 2,

      # Keep observed positive smoking values
      ple0081_h > 0 ~ ple0081_h,

      # Set other invalid cases to NA
      TRUE ~ NA_real_
    ),

    # Create smoking dummy (1 = Smoker, 0 = Non-Smoker)
    smoking_dummy = case_when(
      ple0081_h <= 1.5 ~ 1,
      ple0081_h > 1.5 ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()


# Step 3: Remove Remaining Invalid or Negative Values
main <- main %>%
  filter(!is.na(smoking_dummy))

# Step 4: Descriptive Statistics
# Smoking Dummy
smoking_stats_all <- main %>%
  summarise(
    Mean_Smoking = mean(smoking_dummy, na.rm = TRUE),
    Count = n()
  )

smoking_stats_imputed <- main %>%
  filter(smoking_imputed == 1) %>%
  summarise(
    Mean_Smoking = mean(smoking_dummy, na.rm = TRUE),
    Count = n()
  )

smoking_stats_non_imputed <- main %>%
  filter(smoking_imputed == 0) %>%
  summarise(
    Mean_Smoking = mean(smoking_dummy, na.rm = TRUE),
    Count = n()
  )

print("Descriptive Statistics for Smoking (All Rows):")
print(smoking_stats_all)

print("Descriptive Statistics for Smoking (Imputed Rows):")
print(smoking_stats_imputed)

print("Descriptive Statistics for Smoking (Non-Imputed Rows):")
print(smoking_stats_non_imputed)


# Step 6: Visualizations
# Smoking Dummy
ggplot(main, aes(x = factor(smoking_dummy))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Smoking Status - All Rows", x = "Smoking Status (1 = Yes, 0 = No)", y = "Count") +
  theme_minimal()

ggplot(main %>% filter(smoking_imputed == 1), aes(x = factor(smoking_dummy))) +
  geom_bar(fill = "orange", color = "black") +
  labs(title = "Smoking Status - Imputed Rows", x = "Smoking Status (1 = Yes, 0 = No)", y = "Count") +
  theme_minimal()

ggplot(main %>% filter(smoking_imputed == 0), aes(x = factor(smoking_dummy))) +
  geom_bar(fill = "green", color = "black") +
  labs(title = "Smoking Status - Non-Imputed Rows", x = "Smoking Status (1 = Yes, 0 = No)", y = "Count") +
  theme_minimal()


```

## 3.2.4 Diet: Special Potential Specification because surveyed every second year from 2004-2014
Kick this out because it is way too little observations


# 4. Control Variables
## 4.1 Worries about job security, financial situation, and economic situation 
```{r}
# Create variables for worries about job security, financial situation, and economic situation
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    plh0042_refused = sum(plh0042 == -1, na.rm = TRUE),
    plh0042_NA = sum(plh0042 %in% c(-5, -6), na.rm = TRUE),
    plh0042_not_applicable = sum(plh0042 == -2, na.rm = TRUE),
    plh0033_refused = sum(plh0033 == -1, na.rm = TRUE),
    plh0033_invalid = sum(plh0033 == -2, na.rm = TRUE),
    plh0032_NA = sum(plh0032  == -5, na.rm = TRUE),
    plh0032_invalid = sum(plh0032  == -1, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Lag/Lead Imputation (Max 2 Years)
main <- main %>%
  filter(plh0033 != -1 & plh0042 != -1 & plh0032  != -1 & plh0033 != -2 & plh0032 != -2) %>%
  arrange(pid, syear) %>%  # Ensure data is ordered by pid and syear
  group_by(pid) %>%
  mutate(
    # Impute -5 or -8 for `plh0042` with max 2-year lag/lead
    worries_job_imputed = if_else(plh0042 %in% c(-5, -6), 1, 0),
    plh0042 = case_when(
      plh0042 %in% c(-5, -8) ~ {
        lag_val <- if_else(syear - lag(syear) <= 2, lag(plh0042), NA_real_)
        lead_val <- if_else(lead(syear) - syear <= 2, lead(plh0042), NA_real_)
        if_else(!is.na(lag_val) & !is.na(lead_val), (lag_val + lead_val) / 2, coalesce(lag_val, lead_val))
      },
      plh0042 == -2 ~ 3,
      TRUE ~ plh0042
    ),
    # not necessary for plh0033 because no remaining missings
    # Repeat for `plh0032`
    worries_economic_imputed = if_else(plh0032 == -5, 1, 0),
    plh0032 = case_when(
      plh0032 == -5 ~ {
        lag_val <- if_else(syear - lag(syear) <= 2, lag(plh0032), NA_real_)
        lead_val <- if_else(lead(syear) - syear <= 2, lead(plh0032), NA_real_)
        if_else(!is.na(lag_val) & !is.na(lead_val), (lag_val + lead_val) / 2, coalesce(lag_val, lead_val))
      },
      TRUE ~ plh0032
    )
  ) %>%
  ungroup() 

# Step 3: Create Categorical and Dummy Variables
main <- main %>%
  mutate(
    worr_job_categorical = case_when(plh0042 == 1 ~ 3, # no worries
                                     plh0042 == 3 ~ 1, # large worries
                                     TRUE ~ plh0042),
    worr_financial_categorical = case_when(plh0033 == 1 ~ 3, # no worries
                                          plh0033 == 3 ~ 1, # large worries
                                          TRUE ~ plh0033),
    worr_economic_categorical = case_when(plh0032 == 1 ~ 3, # no worries
                                          plh0032 == 3 ~ 1, # large worries
                                          TRUE ~ plh0032),
    worr_job_dummy = ifelse(worr_job_categorical >= 2, 1, 0),
    worr_financial_dummy = ifelse(worr_financial_categorical >= 2, 1, 0),
    worr_economic_dummy = ifelse(worr_economic_categorical >= 2, 1, 0)
  )

main <- main %>%
  filter(!is.na(worr_job_categorical) & !is.na(worr_financial_categorical)& !is.na(worr_economic_categorical)) # XXX observations dropped

```
# Continuation of code
```{r}
# Define descriptive_stats function
descriptive_stats <- function(data, condition) {
  data %>%
    filter(!!rlang::enquo(condition)) %>%  # Use enquo for capturing condition
    summarise(
      Job_Worries_Mean = mean(as.numeric(worr_job_dummy), na.rm = TRUE),
      Financial_Worries_Mean = mean(as.numeric(worr_financial_dummy), na.rm = TRUE),
      Economic_Worries_Mean = mean(as.numeric(worr_economic_dummy), na.rm = TRUE),
      Count = n()
    )
}
# Calculate statistics for all rows, imputed rows, and non-imputed rows
all_stats <- descriptive_stats(main, TRUE)
imputed_stats <- descriptive_stats(main, worries_job_imputed == 1 | worries_economic_imputed == 1)
non_imputed_stats <- descriptive_stats(main, worries_job_imputed == 0 & worries_economic_imputed == 0)

print("Descriptive Statistics for All Rows:")
print(all_stats)
print("Descriptive Statistics for Imputed Rows:")
print(imputed_stats)
print("Descriptive Statistics for Non-Imputed Rows:")
print(non_imputed_stats)

# Step 5: Visualizations
# Drop rows with NAs in worry-related columns
main <- main %>%
  drop_na(worr_job_categorical, worr_financial_categorical, worr_economic_categorical)

# Distribution of Worry Levels - All Rows
main %>%
  select(worr_job_categorical, worr_financial_categorical, worr_economic_categorical) %>%
  pivot_longer(cols = everything(), names_to = "Worry_Type", values_to = "Worry_Level") %>%
  ggplot(aes(x = Worry_Level, fill = Worry_Type)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Distribution of Worry Levels - All Rows",
    x = "Worry Level",
    y = "Count",
    fill = "Worry Type"
  ) +
  theme_minimal()

# Imputed Rows
main %>%
  filter(worries_job_imputed == 1 | worries_economic_imputed == 1) %>%
  pivot_longer(cols = c(worr_job_categorical, worr_economic_categorical), 
               names_to = "Worry_Type", values_to = "Worry_Level") %>%
  ggplot(aes(x = Worry_Level, fill = Worry_Type)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Distribution of Worry Levels - Imputed Rows",
    x = "Worry Level",
    y = "Count",
    fill = "Worry Type"
  ) +
  theme_minimal()

# Non-Imputed Rows
main %>%
  filter(worries_job_imputed == 0 & worries_economic_imputed == 0) %>%
  pivot_longer(cols = c(worr_job_categorical, worr_economic_categorical), 
               names_to = "Worry_Type", values_to = "Worry_Level") %>%
  ggplot(aes(x = Worry_Level, fill = Worry_Type)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Distribution of Worry Levels - Non-Imputed Rows",
    x = "Worry Level",
    y = "Count",
    fill = "Worry Type"
  ) +
  theme_minimal()

```

## 4.2 Tenure
```{r}
# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    refused = sum(pgerwzeit %in% c(-1, -3), na.rm = TRUE),
    not_applicable = sum(pgerwzeit == -2, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Clean Tenure
main <- main %>%
  filter(pgerwzeit != -1 & pgerwzeit != -3) %>%
  mutate(tenure = case_when(pgerwzeit == -2 ~ 0,
                            TRUE ~ pgerwzeit))

# Step 3: Visualizations
# All Rows
ggplot(main, aes(x = tenure)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Tenure - All Rows",
    x = "Tenure (Years)",
    y = "Count"
  ) +
  theme_minimal()

```

## 4.3 Net Income and Working time per week
```{r}
# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    refused = sum(pglabnet %in% c(-1, -3), na.rm = TRUE),
    not_applicable = sum(pglabnet == -2, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Transform `pglabnet` into `net_income`
main <- main %>%
  mutate(net_income = case_when(pglabnet == -2 ~ 0,
                            TRUE ~ pglabnet))


# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    refused = sum(pgtatzeit %in% c(-1, -3), na.rm = TRUE),
    not_applicable = sum(pgtatzeit == -2, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Transform `pgtatzeit` into `work_time_weekly`
main <- main %>%
  filter(pgtatzeit != -1 & pgtatzeit != -3) %>% 
  mutate(
    work_time_weekly = case_when(pgtatzeit == -2 ~ 0,
                            TRUE ~ pgtatzeit))

# Summary Statistics for Net Income
# All rows
net_income_all <- main %>%
  summarise(
    Mean = mean(net_income, na.rm = TRUE),
    Median = median(net_income, na.rm = TRUE),
    Min = min(net_income, na.rm = TRUE),
    Max = max(net_income, na.rm = TRUE),
    Count = sum(!is.na(net_income))
  )


# Summary Statistics for Work Time Weekly
# All rows
work_time_all <- main %>%
  summarise(
    Mean = mean(work_time_weekly, na.rm = TRUE),
    Median = median(work_time_weekly, na.rm = TRUE),
    Min = min(work_time_weekly, na.rm = TRUE),
    Max = max(work_time_weekly, na.rm = TRUE),
    Count = sum(!is.na(work_time_weekly))
  )

# Display Results
print("Net Income Summary Statistics - All Rows:")
print(net_income_all)

print("\nWork Time Weekly Summary Statistics - All Rows:")
print(work_time_all)

# Plot for Net Income - All Rows
ggplot(main, aes(x = net_income)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Net Income Distribution - All Rows",
    x = "Net Income",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Plot for Work Time Weekly - All Rows
ggplot(main, aes(x = work_time_weekly)) +
  geom_histogram(binwidth = 5, fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(
    title = "Work Time Weekly Distribution - All Rows",
    x = "Work Time (Hours/Week)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

## 4.4 Unemployment
```{r}
# Transform total years unemployed (pgexpue)
# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    not_answered = sum(pgexpue == -1 , na.rm = TRUE)) # 1278

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Clean Total Years Unemployment
main <- main %>%
  rename(total_years_unemployment = pgexpue) %>%
  filter(total_years_unemployment != -1)

# Step 5: Visualizations
# All Rows
ggplot(main, aes(x = total_years_unemployment)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Total Years Unemployed - All Rows",
    x = "Total Years Unemployed",
    y = "Count"
  ) +
  theme_minimal()
```

## 4.5 Occupational Status
```{r}
# Transform occupational status (pgstib)
main <- main %>%
  filter(pgstib!=-1) %>%
  mutate(
    occup_status = case_when(
      between(pgstib, 210, 250) ~ "Blue Collar Employee",  # Blue collar emp
      between(pgstib, 510, 550) ~ "White Collar Employee",  # White collar emp
      between(pgstib, 310, 340) ~ "Agricultural Employee", # agriculture emp
      between(pgstib, 410, 413) ~ "Agriculture - Self-Employed", # agriculture self-emp
      between(pgstib, 610, 640) ~ "Public Servant",  # Public servant
      between(pgstib, 420, 433) | pgstib == 560 ~ "White Collar - Self-Employed",  # Self-employed
      pgstib == 12 | pgstib == -2 | pgstib == 10 ~ "Unemployed" ,  # Unemployed
      pgstib == 11 | pgstib == 15 | between(pgstib, 110, 150) ~ "Apprentice",  # Apprentice
      pgstib == 13 ~ "Retired",  # Retired
      TRUE ~ "Other"  # Assign a default category for unmatched values
    )
  )

main <- main %>%
  mutate(unemp_dummy = ifelse(occup_status == "Unemployed", 1, 0))


#Visualisation: Bar Plot
ggplot(main, aes(x = occup_status)) +
  geom_bar(fill = "darkred", color = "black") +
  labs(
    title = "Distribution of Occupational Status",
    x = "Occupational Status",
    y = "Count"
  ) +
  theme_minimal()

```

## 4.6 Gender
```{r}
# Transform sex into male dummy
main <- main %>%
  filter(sex != -3) %>%
  mutate(male = ifelse(sex == 1, 1, 0))

#Visualisation: Pie Chart
main %>%
  group_by(male) %>%
  summarize(count = n()) %>%
  mutate(
    proportion = count / sum(count),
    label = paste0(round(proportion * 100, 1), "%")
  ) %>%
  ggplot(aes(x = "", y = proportion, fill = factor(male, labels = c("Female", "Male")))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(
    title = "Proportion of Male vs. Female",
    fill = "Gender"
  ) +
  theme_minimal() +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5))
```

## 4.7 Migration Background
```{r}
# Transform migration background and create dummy
main <- main %>%
  mutate(germborn = ifelse(migback == 2, 1, 0),
         migback=ifelse(migback >= 2, 1, 0))

#Visualisation: Bar plot
ggplot(main, aes(x = migback)) +
  geom_bar(fill = "cyan", color = "black") +
  labs(
    title = "Distribution of Migration Background",
    x = "Migration Background",
    y = "Count"
  ) +
  theme_minimal()

```

## 4.8 Marital Status
```{r}
# Count the number of NA rows before imputation
na_count_before <- sum(main$pgfamstd_dummy == -5)
print(paste("Number of NA rows in pgfamstd_dummy before imputation:", na_count_before))
  # depending on that make imputation
# Transform marital status and handle NA with lag/lead logic
main <- main %>%
  filter(pgfamstd != -1 & pgfamstd != -3) %>%
  mutate(rel_status = case_when(pgfamstd == 1 | pgfamstd == 7  ~ "married",
                                pgfamstd == 2 | pgfamstd == 8  ~ "separated",
                                pgfamstd == 3 ~ "single",
                                pgfamstd == 4 ~ "divorced",
                                pgfamstd == 5 ~ "widowed",
                                pgfamstd == 6 ~ "spouse abroad",
                                TRUE ~ NA_character_),
         rel_status = factor(rel_status))

sum(is.na(main$rel_status)) 

# Drop the two missings because relationship status is too fluctuating
main <- main %>%
  filter(!is.na(rel_status))


# Visualizations
# All Rows
ggplot(main, aes(x = rel_status)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Relationship Status - All Rows",
    x = "Relationship Status",
    y = "Count"
  ) +
  theme_minimal()

```
## 4.9 Age
```{r}
main <- main %>%
  filter(gebjahr != -1) %>%
  mutate(age = syear - gebjahr)

# Step 5: Visualizations
# All Rows
ggplot(main, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Agge",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()

main <- main %>%
  filter(between(age, 25, 55))

ggplot(main, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Agge",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()

```


## 4.10 Education
```{r}
test <- main %>%
  filter(pgisced11 != -1)

sum(main$pgisced11 == -2)

main <- main %>%
  mutate(educ = as.numeric(pgisced11),
         educ = case_when(educ < 0 ~ NA_real_,
                             TRUE ~ educ))
main <- main %>%
  mutate(educ_imputed = ifelse(is.na(educ), 1,0)) %>%
  arrange(pid, syear) %>%  # Ensure data is sorted by individual and time
  group_by(pid) %>%
  fill(educ, .direction = "down") %>%  # Carry forward values
  fill(educ, .direction = "up") %>% #also carry upward because all observations are older than 25
  ungroup() %>%
  filter(!is.na(educ)) 

# Labels
# [1] Primary education 
#           [2] Lower secondary education 
#           [3] Upper secondary education 
# [4] Post-secondary non-tertiary educati 
#      [5] Short-cycle tertiary education 
#      [6] Bachelor s or equivalent level 
#        [7] Master s or equivalent level 
#        [8] Doctoral or equivalent level 

# All Rows
ggplot(main, aes(x = educ)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Relationship Status - All Rows",
    x = "Relationship Status",
    y = "Count"
  ) +
  theme_minimal()

# Imputed Rows only
main %>%
  filter(educ_imputed == 1) %>%
  ggplot(aes(x = educ)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Relationship Status - Imputed Rows",
    x = "Relationship Status",
    y = "Count"
  ) +
  theme_minimal()

# Non-Imputed Rows only
main %>%
  filter(educ_imputed == 0) %>%
  ggplot(aes(x = educ)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Relationship Status - Not Imputed Rows",
    x = "Relationship Status",
    y = "Count"
  ) +
  theme_minimal()

```


# Final Cleanup: Keeping relevant & cleaned variables
```{r}
print(colnames(main))
# Select only the relevant cleaned variables
main <- main %>%
  select(
    pid, syear, gebjahr, migback, germborn, health, worr_health_categorical,
    total_years_unemployment, worr_health_dummy, health_in_2yrs, health_decline_2yrs,
    life_satisfaction, health_satisfaction, height, weight, BMI, bmi_categorical,
    smoking_dummy, ever_smoked, worr_job_categorical, worr_financial_categorical, worr_economic_categorical, worr_job_dummy, worr_financial_dummy,worr_economic_dummy,
    tenure, net_income, work_time_weekly, occup_status, unemp_dummy, male, rel_status,
    age, educ
  )

# Count missing values for each column
missing_counts <- colSums(is.na(main))

# View the result
missing_counts

# Count values below 0 for each column
below_zero_counts <- colSums(main < 0, na.rm = TRUE)

# View the result
below_zero_counts

main <- main %>%
  filter(!is.na(health_decline_2yrs) & !is.na(health_in_2yrs) & worr_job_categorical > 0 & worr_economic_categorical > 0)

#Display the summary of the dataset
skim(main)
```

# Saving the cleaned dataset
```{r}
# Save the final main dataset as a CSV file
write.csv(main, "SOEP_main_final.csv", row.names = FALSE)
```