---
title: "SOEP Data Preparation"
author: "Salma-Anastasiia-Suzanna"
date: "2024-12-04"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    df_print: paged
---
# Introduction
We are examining the extent to which individuals’ subjective well-being can be predicted by objective factors that determine individual health decline and for which group this declaration bias is particularly large. And in that sense, which predictors (such as age, labor market status, personality traits and hand-grip strength) help improve predictions of individual health declines.

This report documents the steps taken to preprocess and analyze the SOEP data for health-related outcomes: subjective and objective health measures, predictors, and control variables. The transformations include cleaning, imputing missing values, and deriving new variables.

------------------------------------------------------------------------

# 1. Setup

## 1.1 Load Required Packages

```{r}
# Load required packages
my_packages <- c("tidyverse", "haven", "openxlsx", "labelled", "dplyr", "lubridate","ggplot2","skimr","ggcorrplot","patchwork","ggridges", "glmnet", "plm", "magrittr", "caret","waffle","randomForest","rpart","ranger","vip","pdp","rpart.plot")

for (p in my_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, dependencies = TRUE)
  }
  library(p, character.only = TRUE)
}
```

## 1.2 Load the pl Dataset

```{r}
main <- read_dta('/Users/salmahoumane/Documents/R/pl.dta') %>%
  select(
    "pid", "syear", "ple0008", "plh0035", "ple0055", "ple0056", "ple0072", "ple0081_h", "ple0086_v4",
    "ple0080_v2", "ple0086_v1", "ple0086_v4", paste0("ple00", 11:23),
    "ple0004", "ple0005", "ple0095", "plh0182", "plh0171",
    "plh0042", "plh0033", "plh0032", "ple0010_h", "ple0006", "ple0007"
  )
cat("Main dataset dimensions:", dim(main), "\n")
```

## 1.3 Merge with Other Datasets

```{r}
# Merge with `pgen` dataset
data_pgen <- read_dta('/Users/salmahoumane/Documents/R/pgen.dta') %>%
  select("pid", "syear", "pgerwzeit", "pgemplst", "pglabnet", "pgtatzeit", "pgexpue", "pgstib","pgfamstd", "pgisced11")
cat("PGEN dataset dimensions:", dim(data_pgen), "\n")
main_before_merge <- dim(main)  # Store dimensions before merge
main <- merge(main, data_pgen, by = c("pid", "syear"), all.x = TRUE)
cat("After merging with pgen: Rows =", nrow(main), "Cols =", ncol(main), "\n")
cat("Same number of rows then the main dataset with 7 new rows -> Merge successful")

# Check for duplicate rows introduced during merge
duplicates <- main %>%
  group_by(pid, syear) %>%
  summarise(count = n()) %>%
  filter(count > 1)

if (nrow(duplicates) > 0) {
  cat("Warning: Duplicates detected after merging with pgen!\n")
  print(duplicates)
} else {
  cat("No duplicates found after merging with pgen.\n")
}

# Merge with `ppathl` dataset
data_ppathl <- read_dta('/Users/salmahoumane/Documents/R/ppathl.dta') %>%
  select("pid", "hid", "syear", "psample", "gebjahr", "sex", "migback", "germborn")
cat("PPATHL dataset dimensions:", dim(data_ppathl), "\n")
main_before_ppathl <- dim(main)  # Store dimensions before second merge

main <- left_join(main, data_ppathl, by = c("pid", "syear"))
cat("After merging with ppathl: Rows =", nrow(main), "Cols =", ncol(main), "\n")
cat("Same number of rows then the main dataset with 6 new rows -> Merge successful")

# Final snapshot of dimensions
cat("Final dataset dimensions after all merges:", dim(main), "\n")
```

## 1.4 Main filtering: Analysis post 1999

```{r}
# Drop observations prior to 1999
main <- main %>%
  filter(syear >= 1999)
cat("Dimensions:", dim(main), "\n")
```

# 2. Outcome Variables: Subjective Health

## 2.1 Health

```{r}
# Clean and rename the `health` variable
main <- main %>%
  rename(health = ple0008) %>%
  filter(health != -1) %>%  # Filter valid health values -> 398 observations dropped
    mutate(health = case_when(health == 5 ~ 1, # Reverse the scale: 1 -> 5, 5 -> 1
                            health == 4 ~ 2,
                            health == 3 ~3,
                            health == 2 ~ 4,
                            health == 1 ~ 5))

sum(is.na(main$health))  # no missings

# Visualisation: Bar plot
ggplot(main, aes(x = factor(health))) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Self-Assessed Health",
    x = "Health Category (1 = Poor, 5 = Excellent)",
    y = "Count"
  ) +
  theme_minimal()
```

## 2.2 Worries About Health

```{r}
# Create `worr_health_categorical` and `worr_health_dummy`
main <- main %>%
  rename(worr_health_categorical = plh0035) %>%
  filter(worr_health_categorical != -1 & worr_health_categorical!= -4) %>% # drop those who refused to answer -> 1287 obs
  mutate(
    worr_health_categorical = case_when(
      worr_health_categorical < 0 ~ NA_real_,
      TRUE ~ worr_health_categorical
    )
  ) %>%
  group_by(pid) %>%
  mutate(
    worr_health_categorical = if_else(
      syear %in% c(2011, 2012, 2013) & is.na(worr_health_categorical), #impute the years 2011-2013 because 11000 observations werent asked this question in these years
      floor(
        mean(
          worr_health_categorical[syear %in% c(2010, 2014)],
          na.rm = TRUE
        )
      ),
      worr_health_categorical
    )
  ) %>%
  ungroup() %>%
  drop_na(worr_health_categorical) %>% # 1032
  mutate(worr_health_categorical = case_when(
                            worr_health_categorical == 3 ~ 1, # Reverse the scale: 1 = no worries
                            worr_health_categorical == 2 ~ 2,
                            worr_health_categorical == 1 ~ 3), # 3 = lots of worries
    worr_health_dummy = if_else(
      worr_health_categorical >= 2, 1, 0
    )
  )

# Visualisation: Bar plot (categorical)
ggplot(main, aes(x = worr_health_categorical)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Distribution of Worries About Health",
    x = "Worries About Health",
    y = "Count"
  ) +
  theme_minimal()
# Visualisation: Pie Char (dummy)
main %>%
  group_by(worr_health_dummy) %>%
  summarize(count = n()) %>%
  mutate(
    proportion = count / sum(count),
    label = paste0(round(proportion * 100, 1), "%")
  ) %>%
  ggplot(aes(x = "", y = proportion, fill = factor(worr_health_dummy))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(
    title = "Proportion of Health Worry (Dummy)",
    fill = "Health Worry (Dummy)"
  ) +
  theme_minimal() +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5))
```

## 2.3 Health Trends

```{r}
# Create health decline dummy for 2 years into the future
main <- main %>%
  arrange(pid, syear) %>%  # Ensure data is sorted by pid and syear
  group_by(pid) %>%
  mutate(
    health_in_2yrs = dplyr::lead(health, n = 2, order_by = syear),  # Use dplyr::lead()
    health_decline_2yrs = case_when(
      !is.na(health_in_2yrs) & health_in_2yrs < health ~ 1,  # Decline
      !is.na(health_in_2yrs) & health_in_2yrs >= health ~ 0,  # No decline
      TRUE ~ NA_real_  # NA for missing data
    )
  ) %>%
  ungroup()

# Check for missing values in health_in_2yrs
sum(is.na(main$health_in_2yrs))  # 81.113
#Visualisation: Line plot
# Line plot: Average health over years
main %>%
  group_by(syear) %>%
  summarize(avg_health = mean(health, na.rm = TRUE)) %>%
  ggplot(aes(x = syear, y = avg_health)) +
  geom_line(color = "darkgreen") +
  geom_point(color = "darkgreen") +
  labs(
    title = "Average Self-Assessed Health Over Years",
    x = "Survey Year",
    y = "Average Health"
  ) +
  theme_minimal()
```

# 3. Predictors
## 3.2 Life Satisfaction and Health Satisfaction
```{r}
# Create life_satisfaction and health_satisfaction variables
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    plh0182_NA = sum(is.na(plh0182)),
    plh0182_not_asked = sum(plh0182 %in% c(-8, -5), na.rm = TRUE),
    plh0182_other_neg = sum(plh0182 < 0 & !(plh0182 %in% c(-8, -5)), na.rm = TRUE),
    plh0171_NA = sum(is.na(plh0171)),
    plh0171_not_asked = sum(plh0171 %in% c(-8, -5), na.rm = TRUE),
    plh0171_other_neg = sum(plh0171 < 0 & !(plh0171 %in% c(-8, -5)), na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Drop those who refused to answer or answered invalidly
main <- main %>%
  filter(!(plh0182 %in% c(-1, -2, -3) | plh0171 %in% c(-1, -2, -3))) # 908 observations

# Step 3: Impute remaining Values for plh0182 and plh0171 with Max 2-Year Lag/Lead
main <- main %>%
  arrange(pid, syear) %>%  # Ensure proper sorting
  group_by(pid) %>%
  mutate(
    # Flag rows that will be imputed
    imputed_flag_life = if_else(plh0182 %in% c(-5, -8) | is.na(plh0182), 1, 0),
    imputed_flag_health = if_else(plh0171 %in% c(-5, -8) | is.na(plh0171), 1, 0),

    # Replace -5 and -8 with NA
    plh0182 = case_when(plh0182 %in% c(-5, -8) ~ NA_real_, TRUE ~ plh0182),
    plh0171 = case_when(plh0171 %in% c(-5, -8) ~ NA_real_, TRUE ~ plh0171),

    # Impute plh0182 (life satisfaction)
    life_satisfaction = if_else(
      is.na(plh0182),
      coalesce(
        if_else(syear - lag(syear) <= 2 & !is.na(lag(plh0182)), lag(plh0182), NA_real_),
        if_else(dplyr::lead(syear) - syear <= 2 & !is.na(dplyr::lead(plh0182)), dplyr::lead(plh0182), NA_real_)
      ),
      plh0182
    ),

    # Impute plh0171 (health satisfaction)
    health_satisfaction = if_else(
      is.na(plh0171),
      coalesce(
        if_else(syear - lag(syear) <= 2 & !is.na(lag(plh0171)), lag(plh0171), NA_real_),
        if_else(dplyr::lead(syear) - syear <= 2 & !is.na(dplyr::lead(plh0171)), dplyr::lead(plh0171), NA_real_)
      ),
      plh0171
    )
  ) %>%
  ungroup()

# Step 3: Breakdown After Imputation
missing_breakdown_after <- main %>%
  summarise(
    life_satisfaction_NA = sum(is.na(life_satisfaction)),
    health_satisfaction_NA = sum(is.na(health_satisfaction))
  )

print("Breakdown of missing values after imputation:")
print(missing_breakdown_after)

# Drop observations that are still missing
main <- main %>%
  filter(!is.na(life_satisfaction) & !is.na(health_satisfaction)) # 681 observations dropped

# Step 4: Descriptive Statistics
# Imputed Rows
imputed_stats <- main %>%
  filter(imputed_flag_life == 1 | imputed_flag_health == 1) %>%
  summarise(
    Life_Satisfaction_Mean = mean(life_satisfaction, na.rm = TRUE),
    Life_Satisfaction_Median = median(life_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Mean = mean(health_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Median = median(health_satisfaction, na.rm = TRUE),
    Count = n()
  )

# Non-Imputed Rows
non_imputed_stats <- main %>%
  filter(imputed_flag_life == 0 & imputed_flag_health == 0) %>%
  summarise(
    Life_Satisfaction_Mean = mean(life_satisfaction, na.rm = TRUE),
    Life_Satisfaction_Median = median(life_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Mean = mean(health_satisfaction, na.rm = TRUE),
    Health_Satisfaction_Median = median(health_satisfaction, na.rm = TRUE),
    Count = n()
  )

print("Descriptive Statistics for Imputed Rows:")
print(imputed_stats)

print("Descriptive Statistics for Non-Imputed Rows:")
print(non_imputed_stats)

# The imputed observations display slighly larger health & life-satisfaction

# Step 5: Visualizations
# All Rows: Life Satisfaction
ggplot(main, aes(x = factor(life_satisfaction))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Life Satisfaction - All Rows",
    x = "Life Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Imputed Rows Only
ggplot(main %>% filter(imputed_flag_life == 1), aes(x = factor(life_satisfaction))) +
  geom_bar(fill = "orange", color = "black") +
  labs(
    title = "Life Satisfaction - Imputed Rows",
    x = "Life Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Non-Imputed Rows Only
ggplot(main %>% filter(imputed_flag_life == 0), aes(x = factor(life_satisfaction))) +
  geom_bar(fill = "green", color = "black") +
  labs(
    title = "Life Satisfaction - Non-Imputed Rows",
    x = "Life Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# All Rows: Health Satisfaction
ggplot(main, aes(x = factor(health_satisfaction))) +
  geom_bar(fill = "coral", color = "black") +
  labs(
    title = "Health Satisfaction - All Rows",
    x = "Health Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Imputed Rows Only
ggplot(main %>% filter(imputed_flag_health == 1), aes(x = factor(health_satisfaction))) +
  geom_bar(fill = "orange", color = "black") +
  labs(
    title = "Health Satisfaction - Imputed Rows",
    x = "Health Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()

# Non-Imputed Rows Only
ggplot(main %>% filter(imputed_flag_health == 0), aes(x = factor(health_satisfaction))) +
  geom_bar(fill = "green", color = "black") +
  labs(
    title = "Health Satisfaction - Non-Imputed Rows",
    x = "Health Satisfaction (0-10)",
    y = "Count"
  ) +
  theme_minimal()
```

## 3.2.2 Height, Weight, BMI
```{r}
# Process height and weight
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    height_NA = sum(is.na(ple0006)),
    height_not_asked = sum(ple0006 %in% c(-8, -5), na.rm = TRUE),
    height_invalid = sum(ple0006 > 245 | (ple0006 < 0 & !(ple0006 %in% c(-8, -5))), na.rm = TRUE),
    weight_NA = sum(is.na(ple0007)),
    weight_not_asked = sum(ple0007 %in% c(-8, -5), na.rm = TRUE),
    weight_invalid = sum(ple0007 < 0 & !(ple0007 %in% c(-8, -5)), na.rm = TRUE)
  )

print("Breakdown of missing and invalid values before imputation:")
print(missing_breakdown_before)

# Step 2: Process and Impute Height with Lag/Lead (<= 2 years)
  # Assumption for Imputation of height: does not change significantly for adults
main <- main %>%
  arrange(pid, syear) %>%
  group_by(pid) %>%
  mutate(
    # Track imputation flag for height
    height_imputed = if_else(ple0006 > 245 | ple0006 <= 0 | is.na(ple0006), 1, 0),
    # Set invalid or missing height values to missing
    height = ifelse(ple0006 < 0 | ple0006 > 245, NA, ple0006),
    # Forward fill missing values
    height = zoo::na.locf(height, na.rm = FALSE),

    # Backward fill missing values
    height = zoo::na.locf(height, na.rm = FALSE, fromLast = TRUE)
  ) %>%
  ungroup()

# Step 3: Process and Impute Weight with Lag/Lead (<= 2 years)
#imputation: The mean weight between the last observation before the missing and the first after. And for the ones that have only one before or one after, the one but only if it is an adjacent year.
main <- main %>%
  filter(!(ple0007 %in% c(-1, -2, -3))) %>%
  group_by(pid) %>%
  arrange(syear, .by_group = TRUE) %>%
  mutate(# Track imputation flag for height
    weight_imputed = if_else(ple0007 <= 0 | is.na(ple0007), 1, 0),
    ple0007 = case_when(ple0007<0 ~ NA_real_,
                             TRUE ~ ple0007), 
    # Create lag and lead weights
    lag_weight = lag(ple0007),
    lead_weight = dplyr::lead(ple0007),
    lag_year = lag(syear),
    lead_year = dplyr::lead(syear),
    
    # Apply imputation rules
    weight = case_when(
      is.na(ple0007) & !is.na(lag_weight) & !is.na(lead_weight) & 
        (syear - lag_year == 1 & lead_year - syear == 1) ~ (lag_weight + lead_weight) / 2, # Mean of valid neighbors
      is.na(ple0007) & !is.na(lag_weight) & (syear - lag_year == 1) ~ lag_weight,         # Use previous value if adjacent
      is.na(ple0007) & !is.na(lead_weight) & (lead_year - syear == 1) ~ lead_weight,     # Use next value if adjacent
      TRUE ~ ple0007  # Keep original if no condition is met
    )
  ) %>%
  select(-lag_weight, -lead_weight, -lag_year, -lead_year) %>%  # Clean up temporary columns
  ungroup()



# Step 4: Track Missing Categories After Imputation
missing_breakdown_after <- main %>%
  summarise(
    height_NA = sum(is.na(height)),
    weight_NA = sum(is.na(weight)),
    height_imputed = sum(height_imputed),
    weight_imputed = sum(weight_imputed)
  )

print("Breakdown of missing values after imputation:")
print(missing_breakdown_after)

main <- main %>%
  filter(!is.na(height) & !is.na(weight))

# Step 5: Calculate BMI and Create Categorical Variable
main <- main %>%
  mutate(
    BMI = as.numeric(weight / ((height / 100) ^ 2)),  # Calculate BMI
    bmi_categorical = case_when(
      BMI < 18.5 ~ 1, #underweight
      BMI >= 18.5 & BMI < 25 ~ 2, #normal
      BMI >= 25 & BMI < 30 ~ 3, #overweight
      BMI >= 30 ~ 4, # obese
      TRUE ~ NA_real_
    )
  )

# Step 6: Descriptive Statistics for Imputed and Non-Imputed Rows
# Descriptive stats for all rows
all_stats <- main %>%
  summarise(
    Mean = mean(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE),
    Min = min(BMI, na.rm = TRUE),
    Max = max(BMI, na.rm = TRUE),
    Count = n()
  )

# Descriptive stats for imputed rows
imputed_stats <- main %>%
  filter(height_imputed == 1 | weight_imputed == 1) %>%
  summarise(
    Mean = mean(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE),
    Min = min(BMI, na.rm = TRUE),
    Max = max(BMI, na.rm = TRUE),
    Count = n()
  )

# Descriptive stats for non-imputed rows
non_imputed_stats <- main %>%
  filter(height_imputed == 0 & weight_imputed == 0) %>%
  summarise(
    Mean = mean(BMI, na.rm = TRUE),
    Median = median(BMI, na.rm = TRUE),
    Q1 = quantile(BMI, 0.25, na.rm = TRUE),
    Q3 = quantile(BMI, 0.75, na.rm = TRUE),
    Min = min(BMI, na.rm = TRUE),
    Max = max(BMI, na.rm = TRUE),
    Count = n()
  )

print("Descriptive Statistics for All Rows:")
print(all_stats)

print("Descriptive Statistics for Imputed Rows:")
print(imputed_stats)

print("Descriptive Statistics for Non-Imputed Rows:")
print(non_imputed_stats)

# Step 7: Visualizations
# Histogram for all rows
ggplot(main, aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black", alpha = 0.7) +
  labs(
    title = "BMI Distribution - All Rows",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()

# Histogram for imputed rows
ggplot(main %>% filter(height_imputed == 1 | weight_imputed == 1), aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "orange", color = "black", alpha = 0.7) +
  labs(
    title = "BMI Distribution - Imputed Rows",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()

# Histogram for non-imputed rows
ggplot(main %>% filter(height_imputed == 0 & weight_imputed == 0), aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "green", color = "black", alpha = 0.7) +
  labs(
    title = "BMI Distribution - Non-Imputed Rows",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()

```


## 3.2.3 Smoking
Number of Cigarettes has too little observations
```{r}
# Create `smoking_dummy`
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    ple0081_h_NA = sum(is.na(ple0081_h)),
    ple0081_h_not_asked = sum(ple0081_h %in% c(-5, -8), na.rm = TRUE),
    ple0081_h_invalid = sum(ple0081_h < 0 & !(ple0081_h %in% c(-8, -5)), na.rm = TRUE)
  )

print("Breakdown of missing and invalid values before imputation:")
print(missing_breakdown_before)

# Step 2: Drop those who refused to answer or answered invalidly
main <- main %>%
  filter(!(ple0081_h %in% c(-1, -3) | ple0086_v4 %in% c(-1, -3))) # 137 observations

# Step 3.1: Impute Smoking Status (ple0081_h) with Lag and Lead (Max 2 Years)
main <- main %>%
  arrange(pid, syear) %>%
  group_by(pid) %>%
  mutate(
    # Flag missing smoking values
    smoking_imputed = if_else(ple0081_h %in% c(-5, -8), 1, 0),

    # Determine if the individual was ever observed to smoke
    ever_smoked = any(ple0081_h == 1, na.rm = TRUE),

    # Impute smoking values
    ple0081_h = case_when(
      # If missing and the person ever smoked, impute using lag/lead within 2 years
      ple0081_h %in% c(-5, -8) & ever_smoked ~ {
        lag_val <- if_else(syear - lag(syear) <= 2, lag(ple0081_h), NA_real_)
        lead_val <- if_else(dplyr::lead(syear) - syear <= 2, dplyr::lead(ple0081_h), NA_real_)
        if_else(
          !is.na(lag_val) & !is.na(lead_val),
          (lag_val + lead_val) / 2,
          coalesce(lag_val, lead_val)
        )
      },
      # If missing and the person never smoked, set to 2
      ple0081_h %in% c(-5, -8) & !ever_smoked ~ 2,

      # Keep observed positive smoking values
      ple0081_h > 0 ~ ple0081_h,

      # Set other invalid cases to NA
      TRUE ~ NA_real_
    ),

    # Create smoking dummy (1 = Smoker, 0 = Non-Smoker)
    smoking_dummy = case_when(
      ple0081_h <= 1.5 ~ 1,
      ple0081_h > 1.5 ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()


# Step 3: Remove Remaining Invalid or Negative Values
main <- main %>%
  filter(!is.na(smoking_dummy))

# Step 4: Descriptive Statistics
# Smoking Dummy
smoking_stats_all <- main %>%
  summarise(
    Mean_Smoking = mean(smoking_dummy, na.rm = TRUE),
    Count = n()
  )

smoking_stats_imputed <- main %>%
  filter(smoking_imputed == 1) %>%
  summarise(
    Mean_Smoking = mean(smoking_dummy, na.rm = TRUE),
    Count = n()
  )

smoking_stats_non_imputed <- main %>%
  filter(smoking_imputed == 0) %>%
  summarise(
    Mean_Smoking = mean(smoking_dummy, na.rm = TRUE),
    Count = n()
  )

print("Descriptive Statistics for Smoking (All Rows):")
print(smoking_stats_all)

print("Descriptive Statistics for Smoking (Imputed Rows):")
print(smoking_stats_imputed)

print("Descriptive Statistics for Smoking (Non-Imputed Rows):")
print(smoking_stats_non_imputed)


# Step 6: Visualizations
# Smoking Dummy
ggplot(main, aes(x = factor(smoking_dummy))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Smoking Status - All Rows", x = "Smoking Status (1 = Yes, 0 = No)", y = "Count") +
  theme_minimal()

ggplot(main %>% filter(smoking_imputed == 1), aes(x = factor(smoking_dummy))) +
  geom_bar(fill = "orange", color = "black") +
  labs(title = "Smoking Status - Imputed Rows", x = "Smoking Status (1 = Yes, 0 = No)", y = "Count") +
  theme_minimal()

ggplot(main %>% filter(smoking_imputed == 0), aes(x = factor(smoking_dummy))) +
  geom_bar(fill = "green", color = "black") +
  labs(title = "Smoking Status - Non-Imputed Rows", x = "Smoking Status (1 = Yes, 0 = No)", y = "Count") +
  theme_minimal()


```

## 3.2.4 Diet: Special Potential Specification because surveyed every second year from 2004-2014
Kick this out because it is way too little observations


# 4. Control Variables
## 4.1 Worries about job security, financial situation, and economic situation 
```{r}
# Create variables for worries about job security, financial situation, and economic situation
# Step 1: Track Missing Categories Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    plh0042_refused = sum(plh0042 == -1, na.rm = TRUE),
    plh0042_NA = sum(plh0042 %in% c(-5, -6), na.rm = TRUE),
    plh0042_not_applicable = sum(plh0042 == -2, na.rm = TRUE),
    plh0033_refused = sum(plh0033 == -1, na.rm = TRUE),
    plh0033_invalid = sum(plh0033 == -2, na.rm = TRUE),
    plh0032_NA = sum(plh0032  == -5, na.rm = TRUE),
    plh0032_invalid = sum(plh0032  == -1, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Lag/Lead Imputation (Max 2 Years)
main <- main %>%
  filter(plh0033 != -1 & plh0042 != -1 & plh0032  != -1 & plh0033 != -2 & plh0032 != -2) %>%
  arrange(pid, syear) %>%  # Ensure data is ordered by pid and syear
  group_by(pid) %>%
  mutate(
    # Impute -5 or -8 for `plh0042` with max 2-year lag/lead
    worries_job_imputed = if_else(plh0042 %in% c(-5, -6), 1, 0),
    plh0042 = case_when(
      plh0042 %in% c(-5, -8) ~ {
        lag_val <- if_else(syear - lag(syear) <= 2, lag(plh0042), NA_real_)
        lead_val <- if_else(dplyr::lead(syear) - syear <= 2, dplyr::lead(plh0042), NA_real_)
        if_else(!is.na(lag_val) & !is.na(lead_val), (lag_val + lead_val) / 2, coalesce(lag_val, lead_val))
      },
      plh0042 == -2 ~ 3,
      TRUE ~ plh0042
    ),
    # not necessary for plh0033 because no remaining missings
    # Repeat for `plh0032`
    worries_economic_imputed = if_else(plh0032 == -5, 1, 0),
    plh0032 = case_when(
      plh0032 == -5 ~ {
        lag_val <- if_else(syear - lag(syear) <= 2, lag(plh0032), NA_real_)
        lead_val <- if_else(dplyr::lead(syear) - syear <= 2, dplyr::lead(plh0032), NA_real_)
        if_else(!is.na(lag_val) & !is.na(lead_val), (lag_val + lead_val) / 2, coalesce(lag_val, lead_val))
      },
      TRUE ~ plh0032
    )
  ) %>%
  ungroup() 

# Step 3: Create Categorical and Dummy Variables
main <- main %>%
  mutate(
    worr_job_categorical = case_when(plh0042 == 1 ~ 3, # no worries
                                     plh0042 == 3 ~ 1, # large worries
                                     TRUE ~ plh0042),
    worr_financial_categorical = case_when(plh0033 == 1 ~ 3, # no worries
                                          plh0033 == 3 ~ 1, # large worries
                                          TRUE ~ plh0033),
    worr_economic_categorical = case_when(plh0032 == 1 ~ 3, # no worries
                                          plh0032 == 3 ~ 1, # large worries
                                          TRUE ~ plh0032),
    worr_job_dummy = ifelse(worr_job_categorical >= 2, 1, 0),
    worr_financial_dummy = ifelse(worr_financial_categorical >= 2, 1, 0),
    worr_economic_dummy = ifelse(worr_economic_categorical >= 2, 1, 0)
  )

main <- main %>%
  filter(!is.na(worr_job_categorical) & !is.na(worr_financial_categorical)& !is.na(worr_economic_categorical)) # XXX observations dropped

```

# Descriptive Statistics

```{r}
# Define descriptive_stats function
descriptive_stats <- function(data, condition) {
  data %>%
    filter(!!rlang::enquo(condition)) %>%  # Use enquo for capturing condition
    summarise(
      Job_Worries_Mean = mean(as.numeric(worr_job_dummy), na.rm = TRUE),
      Financial_Worries_Mean = mean(as.numeric(worr_financial_dummy), na.rm = TRUE),
      Economic_Worries_Mean = mean(as.numeric(worr_economic_dummy), na.rm = TRUE),
      Count = n()
    )
}
# Calculate statistics for all rows, imputed rows, and non-imputed rows
all_stats <- descriptive_stats(main, TRUE)
imputed_stats <- descriptive_stats(main, worries_job_imputed == 1 | worries_economic_imputed == 1)
non_imputed_stats <- descriptive_stats(main, worries_job_imputed == 0 & worries_economic_imputed == 0)

print("Descriptive Statistics for All Rows:")
print(all_stats)
print("Descriptive Statistics for Imputed Rows:")
print(imputed_stats)
print("Descriptive Statistics for Non-Imputed Rows:")
print(non_imputed_stats)

# Step 5: Visualizations
# Drop rows with NAs in worry-related columns
main <- main %>%
  drop_na(worr_job_categorical, worr_financial_categorical, worr_economic_categorical)

# Distribution of Worry Levels - All Rows
main %>%
  select(worr_job_categorical, worr_financial_categorical, worr_economic_categorical) %>%
  pivot_longer(cols = everything(), names_to = "Worry_Type", values_to = "Worry_Level") %>%
  ggplot(aes(x = Worry_Level, fill = Worry_Type)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Distribution of Worry Levels - All Rows",
    x = "Worry Level",
    y = "Count",
    fill = "Worry Type"
  ) +
  theme_minimal()

# Imputed Rows
main %>%
  filter(worries_job_imputed == 1 | worries_economic_imputed == 1) %>%
  pivot_longer(cols = c(worr_job_categorical, worr_economic_categorical), 
               names_to = "Worry_Type", values_to = "Worry_Level") %>%
  ggplot(aes(x = Worry_Level, fill = Worry_Type)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Distribution of Worry Levels - Imputed Rows",
    x = "Worry Level",
    y = "Count",
    fill = "Worry Type"
  ) +
  theme_minimal()

# Non-Imputed Rows
main %>%
  filter(worries_job_imputed == 0 & worries_economic_imputed == 0) %>%
  pivot_longer(cols = c(worr_job_categorical, worr_economic_categorical), 
               names_to = "Worry_Type", values_to = "Worry_Level") %>%
  ggplot(aes(x = Worry_Level, fill = Worry_Type)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Distribution of Worry Levels - Non-Imputed Rows",
    x = "Worry Level",
    y = "Count",
    fill = "Worry Type"
  ) +
  theme_minimal()

```

## 4.2 Tenure
```{r}
# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    refused = sum(pgerwzeit %in% c(-1, -3), na.rm = TRUE),
    not_applicable = sum(pgerwzeit == -2, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Clean Tenure
main <- main %>%
  filter(pgerwzeit != -1 & pgerwzeit != -3) %>%
  mutate(tenure = case_when(pgerwzeit == -2 ~ 0,
                            TRUE ~ pgerwzeit))

# Step 3: Visualizations
# All Rows
ggplot(main, aes(x = tenure)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Tenure - All Rows",
    x = "Tenure (Years)",
    y = "Count"
  ) +
  theme_minimal()

```

## 4.3 Net Income and Working time per week
```{r}
# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    refused = sum(pglabnet %in% c(-1, -3), na.rm = TRUE),
    not_applicable = sum(pglabnet == -2, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Transform `pglabnet` into `net_income`
main <- main %>%
  mutate(net_income = case_when(pglabnet == -2 ~ 0,
                            TRUE ~ pglabnet))


# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    refused = sum(pgtatzeit %in% c(-1, -3), na.rm = TRUE),
    not_applicable = sum(pgtatzeit == -2, na.rm = TRUE)
  )

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Transform `pgtatzeit` into `work_time_weekly`
main <- main %>%
  filter(pgtatzeit != -1 & pgtatzeit != -3) %>% 
  mutate(
    work_time_weekly = case_when(pgtatzeit == -2 ~ 0,
                            TRUE ~ pgtatzeit))

# Summary Statistics for Net Income
# All rows
net_income_all <- main %>%
  summarise(
    Mean = mean(net_income, na.rm = TRUE),
    Median = median(net_income, na.rm = TRUE),
    Min = min(net_income, na.rm = TRUE),
    Max = max(net_income, na.rm = TRUE),
    Count = sum(!is.na(net_income))
  )


# Summary Statistics for Work Time Weekly
# All rows
work_time_all <- main %>%
  summarise(
    Mean = mean(work_time_weekly, na.rm = TRUE),
    Median = median(work_time_weekly, na.rm = TRUE),
    Min = min(work_time_weekly, na.rm = TRUE),
    Max = max(work_time_weekly, na.rm = TRUE),
    Count = sum(!is.na(work_time_weekly))
  )

# Display Results
print("Net Income Summary Statistics - All Rows:")
print(net_income_all)

print("\nWork Time Weekly Summary Statistics - All Rows:")
print(work_time_all)

# Plot for Net Income - All Rows
ggplot(main, aes(x = net_income)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Net Income Distribution - All Rows",
    x = "Net Income",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Plot for Work Time Weekly - All Rows
ggplot(main, aes(x = work_time_weekly)) +
  geom_histogram(binwidth = 5, fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(
    title = "Work Time Weekly Distribution - All Rows",
    x = "Work Time (Hours/Week)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

## 4.4 Unemployment
```{r}
# Transform total years unemployed (pgexpue)
# Step 1: Breakdown Before Imputation
missing_breakdown_before <- main %>%
  summarise(
    not_answered = sum(pgexpue == -1 , na.rm = TRUE)) # 1278

print("Breakdown of missing values before imputation:")
print(missing_breakdown_before)

# Step 2: Clean Total Years Unemployment
main <- main %>%
  rename(total_years_unemployment = pgexpue) %>%
  filter(total_years_unemployment != -1)

# Step 5: Visualizations
# All Rows
ggplot(main, aes(x = total_years_unemployment)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Total Years Unemployed - All Rows",
    x = "Total Years Unemployed",
    y = "Count"
  ) +
  theme_minimal()
```

## 4.5 Occupational Status
```{r}
# Transform occupational status (pgstib)
main <- main %>%
  filter(pgstib != -1) %>%
  mutate(
    occup_status = case_when(
      pgstib >= 210 & pgstib <= 250 ~ "Blue Collar Employee",  # Blue collar emp
      pgstib >= 510 & pgstib <= 550 ~ "White Collar Employee",  # White collar emp
      pgstib >= 310 & pgstib <= 340 ~ "Agricultural Employee", # agriculture emp
      pgstib >= 410 & pgstib <= 413 ~ "Agriculture - Self-Employed", # agriculture self-emp
      pgstib >= 610 & pgstib <= 640 ~ "Public Servant",  # Public servant
      (pgstib >= 420 & pgstib <= 433) | pgstib == 560 ~ "White Collar - Self-Employed",  # Self-employed
      pgstib %in% c(12, -2, 10) ~ "Unemployed",  # Unemployed
      pgstib %in% c(11, 15) | (pgstib >= 110 & pgstib <= 150) ~ "Apprentice",  # Apprentice
      pgstib == 13 ~ "Retired",  # Retired
      TRUE ~ "Other"  # Assign a default category for unmatched values
    )
  )


main <- main %>%
  mutate(unemp_dummy = ifelse(occup_status == "Unemployed", 1, 0))


#Visualisation: Bar Plot
ggplot(main, aes(x = occup_status)) +
  geom_bar(fill = "darkred", color = "black") +
  labs(
    title = "Distribution of Occupational Status",
    x = "Occupational Status",
    y = "Count"
  ) +
  theme_minimal()

```

## 4.6 Gender
```{r}
# Transform sex into male dummy
main <- main %>%
  filter(sex != -3) %>%
  mutate(male = ifelse(sex == 1, 1, 0))

#Visualisation: Pie Chart
main %>%
  group_by(male) %>%
  summarize(count = n()) %>%
  mutate(
    proportion = count / sum(count),
    label = paste0(round(proportion * 100, 1), "%")
  ) %>%
  ggplot(aes(x = "", y = proportion, fill = factor(male, labels = c("Female", "Male")))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(
    title = "Proportion of Male vs. Female",
    fill = "Gender"
  ) +
  theme_minimal() +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5))
```

## 4.7 Migration Background
```{r}
# Transform migration background and create dummy
main <- main %>%
  mutate(germborn = ifelse(migback == 2, 0, 1),
         migback=ifelse(migback >= 2, 1, 0))

#Visualisation: Bar plot
ggplot(main, aes(x = migback)) +
  geom_bar(fill = "cyan", color = "black") +
  labs(
    title = "Distribution of Migration Background",
    x = "Migration Background",
    y = "Count"
  ) +
  theme_minimal()

```

## 4.8 Marital Status
```{r}
# Count the number of NA rows before imputation
na_count_before <- sum(main$pgfamstd_dummy == -5)
print(paste("Number of NA rows in pgfamstd_dummy before imputation:", na_count_before))
  # depending on that make imputation
# Transform marital status and handle NA with lag/lead logic
main <- main %>%
  filter(pgfamstd != -1 & pgfamstd != -3) %>%
  mutate(rel_status = case_when(pgfamstd == 1 | pgfamstd == 7  ~ "married",
                                pgfamstd == 2 | pgfamstd == 8  ~ "separated",
                                pgfamstd == 3 ~ "single",
                                pgfamstd == 4 ~ "divorced",
                                pgfamstd == 5 ~ "widowed",
                                pgfamstd == 6 ~ "spouse abroad",
                                TRUE ~ NA_character_),
         rel_status = factor(rel_status))

sum(is.na(main$rel_status)) 

# Drop the two missings because relationship status is too fluctuating
main <- main %>%
  filter(!is.na(rel_status))


# Visualizations
# All Rows
ggplot(main, aes(x = rel_status)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Relationship Status - All Rows",
    x = "Relationship Status",
    y = "Count"
  ) +
  theme_minimal()

```
## 4.9 Age
```{r}
main <- main %>%
  filter(gebjahr != -1) %>%
  mutate(age = syear - gebjahr)

# Step 5: Visualizations
# All Rows
ggplot(main, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Agge",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()

# Filter for age between 25 and 55
main <- main %>%
  filter(age >= 25 & age <= 55)

ggplot(main, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Agge",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()

```


## 4.10 Education
```{r}
test <- main %>%
  filter(pgisced11 != -1)

sum(main$pgisced11 == -2)

main <- main %>%
  mutate(educ = as.numeric(pgisced11),
         educ = case_when(educ < 0 ~ NA_real_,
                             TRUE ~ educ))
main <- main %>%
  mutate(educ_imputed = ifelse(is.na(educ), 1,0)) %>%
  arrange(pid, syear) %>%  # Ensure data is sorted by individual and time
  group_by(pid) %>%
  fill(educ, .direction = "down") %>%  # Carry forward values
  fill(educ, .direction = "up") %>% #also carry upward because all observations are older than 25
  ungroup() %>%
  filter(!is.na(educ)) 

# Labels
# [1] Primary education 
#           [2] Lower secondary education 
#           [3] Upper secondary education 
# [4] Post-secondary non-tertiary educati 
#      [5] Short-cycle tertiary education 
#      [6] Bachelor s or equivalent level 
#        [7] Master s or equivalent level 
#        [8] Doctoral or equivalent level 

# All Rows
ggplot(main, aes(x = educ)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Education Level - All Rows",
    x = "Education Level",
    y = "Count"
  ) +
  theme_minimal()

# Imputed Rows only
main %>%
  filter(educ_imputed == 1) %>%
  ggplot(aes(x = educ)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Education Level - Imputed Rows",
    x = "Education Level",
    y = "Count"
  ) +
  theme_minimal()

# Non-Imputed Rows only
main %>%
  filter(educ_imputed == 0) %>%
  ggplot(aes(x = educ)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Education Level - Not Imputed Rows",
    x = "Education Level",
    y = "Count"
  ) +
  theme_minimal()

```


# Final Cleanup: Keeping relevant & cleaned variables
```{r}
print(colnames(main))
# Select only the relevant cleaned variables
main <- main %>%
  select(
    pid, syear, gebjahr, migback, germborn, health, worr_health_categorical,
    total_years_unemployment, worr_health_dummy, health_in_2yrs, health_decline_2yrs,
    life_satisfaction, health_satisfaction, height, weight, BMI, bmi_categorical,
    smoking_dummy, ever_smoked, worr_job_categorical, worr_financial_categorical, worr_economic_categorical, worr_job_dummy, worr_financial_dummy,worr_economic_dummy,
    tenure, net_income, work_time_weekly, occup_status, unemp_dummy, male, rel_status,
    age, educ
  )

# Count missing values for each column
missing_counts <- colSums(is.na(main))

# View the result
missing_counts

# Count values below 0 for each column
below_zero_counts <- colSums(main < 0, na.rm = TRUE)

# View the result
below_zero_counts

main <- main %>%
  filter(!is.na(health_decline_2yrs) & !is.na(health_in_2yrs) & worr_job_categorical > 0 & worr_economic_categorical > 0)

#Display the summary of the dataset
skim(main)
```

# Saving the cleaned dataset
```{r}
# Save the final main dataset as a CSV file
write.csv(main, "SOEP_main_final.csv", row.names = FALSE)
```
# Correlation Heatmap
```{r}
# In order to identify key predictors and relationships between variables, we create a correlation heatmap between numerical variables
# Select numeric columns
numeric_data <- main[, sapply(main, is.numeric)]

# Calculate correlation matrix
corr_matrix <- cor(numeric_data, use = "complete.obs")

# Plot correlation heatmap
ggcorrplot(corr_matrix, method = "circle", lab = TRUE) +
  labs(title = "Correlation Heatmap") +
  theme_minimal()

# Set correlations below 0.5 or above -0.5 to NA
filtered_corr <- corr_matrix
filtered_corr[abs(filtered_corr) < 0.5] <- NA


# Define pairs to exclude
exclude_pairs <- list(
  c("worr_job_dummy", "worr_job_categorical"),
  c("worr_job_categorical", "worr_job_dummy"),
  c("worr_economic_dummy", "worr_economic_categorical"),
  c("worr_economic_categorical", "worr_economic_dummy"),
  c("worr_health_dummy", "worr_health_categorical"),
  c("worr_health_categorical", "worr_health_dummy"),
  c("bmi_categorical", "BMI"),
  c("BMI", "bmi_categorical"),
  c("BMI", "weight"),
  c("weight", "BMI"),
  c("germborn", "migback"),
  c("migback", "germborn"),
  c("worr_financial_dummy", "worr_financial_categorical"),
  c("worr_financial_categorical", "worr_financial_dummy"),
  c("age", "gebjahr"),
  c("gebjahr", "age"),
  c("bmi_categorical", "weight"),
  c("weight", "bmi_categorical"),
  c("worr_economic_dummy", "worr_economic_categorical"),
  c("health_satisfaction", "health"),
  c("work_time_weekly", "net_income"),
  c("syear", "pid"),
  c("pid", "syear"),
  c("weight", "height"),
  c("worr_health_categorical", "health"),
  c("work_time_weekly", "unemp_dummy"),
  c("life_satisfaction", "health_satisfaction"),
  c("worr_health_categorical", "health_satisfaction"),
  c("health_in_2yrs", "health"),
  c("male", "height"),
  c("health_in_2yrs", "health_satisfaction")
)

filtered_table <- as.data.frame(as.table(filtered_corr)) %>%
  filter(
    !is.na(Freq),  # Exclude NA correlations
    Var1 != Var2,  # Exclude diagonal elements
    !(paste(Var1, Var2, sep = "_") %in% sapply(exclude_pairs, function(x) paste(x[1], x[2], sep = "_")))  # Exclude specific pairs
  ) %>%
  arrange(desc(abs(Freq)))  # Sort by absolute correlation

# View the filtered table
filtered_table  # Display the table

# Create a bar plot
# Rename variables for prettier and clearer names
filtered_table <- filtered_table %>%
  mutate(
    Variable_Pairs = case_when(
      Var1 == "health" & Var2 == "health_satisfaction" ~ "Health & Health Satisfaction",
      Var1 == "height" & Var2 == "male" ~ "Height & Gender (Male)",
      Var1 == "net_income" & Var2 == "work_time_weekly" ~ "Net Income & Work Time Weekly",
      Var1 == "health" & Var2 == "health_in_2yrs" ~ "Health & Health in 2 years",
      Var1 == "height" & Var2 == "weight" ~ "Height & Weight",
      Var1 == "health_satisfaction" & Var2 == "health_in_2yrs" ~ "Health Satisfaction & Health in 2 years",
      Var1 == "health_satisfaction" & Var2 == "life_satisfaction" ~ "Health Satisfaction & Life Satisfaction",
      Var1 == "health_satisfaction" & Var2 == "worr_health_categorical" ~ "Health Satisfaction & Worrying about Health (Categorical)",
      Var1 == "health" & Var2 == "worr_health_categorical" ~ "Health & Worrying about Health (Categorical)",
      Var1 == "unemp_dummy" & Var2 == "work_time_weekly" ~ "Unemployment (Dummy) & Work Time Weekly",
      TRUE ~ paste(Var1, Var2, sep = " & ") # Default to original pairs if not specified
    )
  )

# Create the bar plot
ggplot(filtered_table, aes(x = reorder(Variable_Pairs, Freq), y = Freq, fill = Freq > 0)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Correlation Plot",
    subtitle = "Visualizing Correlations Across Variable Pairs"
  ) +
  scale_fill_manual(
    values = c("#FF0000", "#163E64"),  # Red for negative, dark blue for positive
    guide = "none"  # Remove the legend
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),  # Remove y-axis title
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis text
    axis.text.y = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis text

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

#Save Graph
ggsave("correlation_plot.png", plot = last_plot(), width = 16, height = 8, dpi = 300)

```

# Gender Disparities
```{r}
# In order to highlight gender disparities in health perceptions, we create these graphs per gender
# Group data by pid and gender (male), and summarize
main_unique <- main %>%
  group_by(pid, male) %>%
  summarize(
    health = mean(health, na.rm = TRUE),
    health_decline_2yrs = mean(health_decline_2yrs, na.rm = TRUE),
    worr_health_dummy = mean(worr_health_dummy, na.rm = TRUE),
    health_satisfaction = mean(health_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )

# Health  per Gender: Violin Plot
ggplot(main_unique, aes(x = factor(male), y = health, fill = factor(male))) +
  geom_violin(trim = FALSE, alpha = 0.7, color = "black") +  # Add transparency to the violins
  stat_summary(fun = "mean", geom = "point", shape = 16, size = 3, color = "black") +  # Add mean points
  geom_text(stat = "summary", fun = "mean", aes(label = round(after_stat(y), 2)), vjust = -0.5) +  # Add mean labels
  scale_x_discrete(labels = c("0" = "Female", "1" = "Male")) +  # Update x-axis labels
  scale_fill_manual(values = c("#FF0000", "#163E64")) +  # Red for Female, Dark Blue for Male
  labs(
    title = "Health Distribution by Gender",
    subtitle = "Violin Plot Showing Health Scores",
    x = "Gender",
    y = "Health"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
#Save Graph
ggsave("gender_health.png", plot = last_plot(), dpi = 300)

# Graph for health_decline_2yrs per Gender: Histogram
ggplot(main_unique, aes(x = factor(male), y = health_decline_2yrs, fill = factor(male))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and define outliers
  scale_fill_manual(values = c("#FF0000", "#163E64"), labels = c("Female", "Male"), name = "Gender") +  # Custom fill colors
  scale_x_discrete(labels = c("0" = "Female", "1" = "Male")) +  # Update x-axis labels
  labs(
    title = "Health Decline in 2 Years by Gender",
    subtitle = "Boxplot Comparison of Health Decline Across Genders",
    x = "Gender",
    y = "Health Decline"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"
  )
#Save Graph
ggsave("gender_health_decline.png", plot = last_plot(), dpi = 300)


# Graph for worr_health_dummy per Gender: Histogram
ggplot(main_unique, aes(x = factor(male), y = worr_health_dummy, fill = factor(male))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and define outliers
  scale_fill_manual(values = c("#FF0000", "#163E64"), labels = c("Female", "Male"), name = "Gender") +  # Custom fill colors
  scale_x_discrete(labels = c("0" = "Female", "1" = "Male")) +  # Update x-axis labels
  labs(
    title = "Health Worries by Gender",
    subtitle = "Boxplot of Health Worries for Females and Males",
    x = "Gender",
    y = "Health Worries"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "None"
  )


#Save Graph
ggsave("gender_worr_health.png", plot = last_plot(), dpi = 300)

# Graph for health_satisfaction per Gender: Density plot
ggplot(main_unique, aes(x = health_satisfaction, fill = factor(male), color = factor(male))) +
  geom_density(alpha = 0.5, size = 1) +  # Add transparency and adjust line thickness
  scale_fill_manual(values = c("#FF0000", "#163E64"), labels = c("Female", "Male"), name = "Gender") +  # Custom fill colors
  scale_color_manual(values = c("#FF0000", "#163E64"), labels = c("Female", "Male"), name = "Gender") +  # Custom line colors
  labs(
    title = "Density of Health Satisfaction by Gender",
    subtitle = "Density Plot Comparing Female and Male Responses",
    x = "Health Satisfaction",
    y = "Density",
    fill = "Gender",
    color = "Gender"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"
  )
#Save Graph
ggsave("gender_health_satisfaction.png", plot = last_plot(), dpi = 300)

```
We notice that gender differences are insignificant on health variables. We, therefore, do not expect gender to be a strong predictor.


# Educational Level Disparities

```{r}

# Group data by pid and educ, and summarize
main_educ <- main %>%
  group_by(pid, educ) %>%
  summarize(
    health = mean(health, na.rm = TRUE),
    health_decline_2yrs = mean(health_decline_2yrs, na.rm = TRUE),
    worr_health_dummy = mean(worr_health_dummy, na.rm = TRUE),
    health_satisfaction = mean(health_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )
# Count of individuals per Educ
ggplot(main_educ %>% filter(educ != 0), aes(x = factor(educ), fill = factor(educ))) +
  geom_bar(alpha = 0.7, color = "black") +
  scale_x_discrete(
    labels = c(
      "1" = "Primary",
      "2" = "Lower Secondary",
      "3" = "Upper Secondary",
      "4" = "Post-Secondary",
      "5" = "Short-Cycle Tertiary",
      "6" = "Bachelor's",
      "7" = "Master's",
      "8" = "Doctoral"
    ),
    limits = c("1", "2", "3", "4", "5", "6", "7", "8") # Exclude 0 explicitly
  ) +
  scale_fill_brewer(palette = "RdBu") +  
  labs(
    title = "Distribution of Individuals by Educational Level",
    x = "Educational Level",
    y = "Count",
    fill = "Educational Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"
  )

#Save Graph
ggsave("educ.png", plot = last_plot(), dpi = 300)

# Health  per Educ: 
ggplot(main_educ %>% filter(educ != 0), aes(x = health, fill = factor(educ))) +
  geom_density(alpha = 0.7, color = "black") +
  facet_wrap(~ factor(educ), ncol = 2, labeller = as_labeller(c(
    "1" = "Primary",
    "2" = "Lower Secondary",
    "3" = "Upper Secondary",
    "4" = "Post-Secondary",
    "5" = "Short-Cycle Tertiary",
    "6" = "Bachelor's",
    "7" = "Master's",
    "8" = "Doctoral"
  ))) +
  scale_fill_brewer(palette = "RdBu", guide = "none") +  # Use diverging red/blue palette
  labs(
    title = "Density of Health by Educational Level",
    x = "Health",
    y = "Density",
    fill = "Educational Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_blank(),  # Dark blue y-axis labels

    # Facet title styling
    strip.text = element_text(size = 12, face = "bold", color = "#1E2B4F", margin = ggplot2::margin(b = 5)),  # Style facet labels
    
    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
  )

#Save Graph
ggsave("educ_health.png", plot = last_plot(), dpi = 300)

# Graph for health_decline_2yrs per educ: 
ggplot(main_educ %>% filter(educ != 0), aes(x = factor(educ), y = health_decline_2yrs, fill = factor(educ))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black border
  scale_x_discrete(
    labels = c(
      "1" = "Primary",
      "2" = "Lower Secondary",
      "3" = "Upper Secondary",
      "4" = "Post-Secondary",
      "5" = "Short-Cycle Tertiary",
      "6" = "Bachelor's",
      "7" = "Master's",
      "8" = "Doctoral"
    )
  ) +
  scale_fill_brewer(palette = "RdBu", guide = "none") +  # Use Blues palette
  labs(
    title = "Health Decline in 2 Years by Educational Level",
    subtitle = "Boxplot of Health Decline Across Education Levels",
    x = "Educational Level",
    y = "Health Decline",
    fill = "Educational Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

#Save Graph
ggsave("educ_health_decline.png", plot = last_plot(), dpi = 300)

# Graph for worr_health_dummy per educ:
ggplot(main_educ %>% filter(educ != 0), aes(x = factor(educ), y = worr_health_dummy, fill = factor(educ))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black borders
  scale_x_discrete(
    labels = c(
      "1" = "Primary",
      "2" = "Lower Secondary",
      "3" = "Upper Secondary",
      "4" = "Post-Secondary",
      "5" = "Short-Cycle Tertiary",
      "6" = "Bachelor's",
      "7" = "Master's",
      "8" = "Doctoral"
    )
  ) +
  scale_fill_brewer(palette = "RdBu", guide = "none") +  # Use Blues palette
  labs(
    title = "Health Worries by Educational Level",
    subtitle = "Boxplot of Health Worries Across Education Levels",
    x = "Educational Level",
    y = "Health Worries",
    fill = "Educational Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )


#Save Graph
ggsave("educ_worr_health.png", plot = last_plot(), dpi = 300)

# Graph for health_satisfaction per educ: 
ggplot(main_educ %>% filter(educ != 0), aes(x = factor(educ), y = health_satisfaction, fill = factor(educ))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black border
  scale_x_discrete(
    labels = c(
      "1" = "Primary",
      "2" = "Lower Secondary",
      "3" = "Upper Secondary",
      "4" = "Post-Secondary",
      "5" = "Short-Cycle Tertiary",
      "6" = "Bachelor's",
      "7" = "Master's",
      "8" = "Doctoral"
    )
  ) +
  scale_fill_brewer(palette = "RdBu", guide = "none") +  
  labs(
    title = "Health Satisfaction by Educational Level",
    subtitle = "Boxplot of Health Satisfaction Across Education Levels",
    x = "Educational Level",
    y = "Health Satisfaction",
    fill = "Educational Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
#Save Graph
ggsave("educ_health_satisfaction.png", plot = last_plot(), dpi = 300)


```
Our dataset mostly contains individuals with upper secondary education levels, with only a few primary school level and doctoral level. We expect those 2 groups to have a high variance.

In all graphs, we notice the same results: positive health perception as well as health satisfaction increases with education, and negative health perception through worrying or health decline decreases with education. We expect education to be correlated with income: people with higher education levels get higher salaries so the effects might be linked.


# Income Level Disparities

There are 3 individuals with income over 40k, we will be dropping them

```{r}
# Aggregate data to calculate mean net income for each pid
cleaned_main <- main %>%
  group_by(pid) %>%
  summarize(net_income = mean(net_income, na.rm = TRUE), .groups = "drop") %>%
  filter(net_income <= 40000) # Remove rows where mean income > 40000

# Distribution of income variable
ggplot(cleaned_main, aes(x = net_income)) +
  geom_histogram(binwidth = 500, fill = "#163E64", color = "black", alpha = 0.7) +  # Use dark blue fill and black borders
  labs(
    title = "Histogram of Net Income Distribution",
    subtitle = "Distribution of Net Income Across Individuals",
    x = "Net Income",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space on the right and top/bottom
  )
ggsave("income.png", plot = last_plot(), dpi = 300)

# Group data by pid and educ, and summarize
income_data <- main %>%
  filter(net_income < 40000) %>%
  group_by(pid, net_income) %>%
  summarize(
    health = mean(health, na.rm = TRUE),
    health_decline_2yrs = mean(health_decline_2yrs, na.rm = TRUE),
    worr_health_dummy = mean(worr_health_dummy, na.rm = TRUE),
    health_satisfaction = mean(health_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )

# Health  per Income: 
ggplot(income_data, aes(x = net_income, y = health)) +
    geom_smooth(
    aes(group = 1),  # Ensure a single LOESS line for all bins
    method = "loess",
    color = "#F00000",  # Red for LOESS line
    size = 1.2,  
    se = FALSE, # No confidence interval shading
    span = 1 
  ) +
  stat_summary_bin(fun = "mean", bins = 30, geom = "point", color = "#163E64", size = 2) +  
  labs(
    title = "Health vs. Income (Binned Scatter Plot)",
    x = "Net Income",
    y = "Average Health"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space on the right and top/bottom
  )

ggsave("income_health.png", plot = last_plot(), dpi = 300)


# Graph for health_decline_2yrs per income: 
income_data <- income_data %>%
  mutate(net_income_bin = cut(net_income, breaks = 10)) # Create income bins
# Reformat income bins into human-readable labels
income_data <- income_data %>%
  mutate(
    net_income_bin = cut(
      net_income, 
      breaks = 10, 
      labels = c(
        "< 2.5K", 
        "2.5K - 5K", 
        "5K - 7.5K", 
        "7.5K - 10K", 
        "10K - 12.5K", 
        "12.5K - 15K", 
        "15K - 17.5K", 
        "17.5K - 20K", 
        "20K - 22.5K", 
        "> 22.5K"
      )
    )
  )

ggplot(income_data, aes(x = net_income_bin, y = health_decline_2yrs, fill = net_income_bin)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black border
  scale_fill_brewer(palette = "RdBu", guide = "none") + 
  labs(
    title = "Health Decline by Income (Binned)",
    x = "Net Income (Binned)",
    y = "Health Decline"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

ggsave("income_health_decline.png", plot = last_plot(), dpi = 300)


# Graph for worr_health_dummy per income:
ggplot(income_data, aes(x = worr_health_dummy, fill = net_income_bin)) +
  geom_density(alpha = 0.5, color = "black", size = 0.5) +  # Add transparency and black outline
  scale_fill_brewer(palette = "RdBu", name = "Net Income (Binned)") +
  labs(
    title = "Density of Health Worries by Income",
    x = "Health Worries",
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "right",  # Move legend to the top
    legend.text = element_text(size = 10, color = "#1E2B4F"),  # Dark blue legend text
    legend.title = element_text(size = 12, face = "bold", color = "#1E2B4F")  # Bold and dark blue legend title
  )

ggsave("income_worr_health.png", plot = last_plot(), dpi = 300)

# Group data by income bins and calculate average health satisfaction
heatmap_data <- income_data %>%
  mutate(net_income_bin = cut(
    net_income, 
    breaks = 10, 
    labels = c(
      "< 2.5K", 
      "2.5K - 5K", 
      "5K - 7.5K", 
      "7.5K - 10K", 
      "10K - 12.5K", 
      "12.5K - 15K", 
      "15K - 17.5K", 
      "17.5K - 20K", 
      "20K - 22.5K", 
      "> 22.5K"
    )
  )) %>%
  group_by(net_income_bin) %>%
  summarize(avg_health_satisfaction = mean(health_satisfaction, na.rm = TRUE), .groups = "drop")

# Create the heatmap
ggplot(heatmap_data, aes(x = net_income_bin, y = 1, fill = avg_health_satisfaction)) +
  geom_tile(color = "white") +  # Add white borders for separation
  scale_fill_gradient2(
    low = "#67001F",  # Deep Red
    mid = "#F7F7F7",  # Neutral White
    high = "#053061", # Deep Blue
    midpoint = mean(heatmap_data$avg_health_satisfaction, na.rm = TRUE)  # Center the gradient around the mean
  ) +
  labs(
    title = "Average Health Satisfaction by Income Group",
    x = "Net Income (Binned)",
    y = "Health Satisfaction",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_blank(),  # Remove y-axis labels (since it’s a single-row heatmap)
    axis.ticks.y = element_blank(),  # Remove y-axis ticks
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"), 

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 70, b = 20, l = 50),  # Add more space around the plot
    
    # Legend styling
    legend.position = "right",  # Move legend to the top
  )

ggsave("income_health_satisfaction.png", plot = last_plot(), dpi = 300)

```
Most of the participants in this dataset have  a net income lower than 5000 euros and we see that for those, the higher the income, the higher the average health as we have a positively sloped line. However, for people with higher than 5000 euros income, we see a high variance, with some outliers therefore the negative trend cannot be confirmed.


# Box Plot of Net Income by Education Level

```{r}
ggplot(main, aes(x = factor(educ), y = net_income, fill = factor(educ))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +
  scale_x_discrete(
    labels = c(
      "1" = "Primary",
      "2" = "Lower Secondary",
      "3" = "Upper Secondary",
      "4" = "Post-Secondary",
      "5" = "Short-Cycle Tertiary",
      "6" = "Bachelor's",
      "7" = "Master's",
      "8" = "Doctoral"
    )
  ) +
  scale_fill_brewer(palette = "RdBu", guide = "none") +  
  labs(
    title = "Net Income by Education Level",
    x = "Education Level",
    y = "Net Income",
    fill = "Education Level"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space on the right and top/bottom
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
ggsave("income_educ.png", plot = last_plot(), dpi = 300)
```
There is not a significant relationship between education level and net income. 

# Age Disparities

```{r}
# Distribution of age variable
ggplot(main, aes(x = age)) +
  geom_histogram(aes(y = ..density..), binwidth = 5, fill = "#A6CEE3", color = "black", alpha = 0.7) + 
  geom_density(color = "#1E2B4F", size = 1) + 
  labs(
    title = "Combined Histogram and Density Plot of Age Distribution",
    x = "Age",
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space around the plot
  )

ggsave("age.png", plot = last_plot(), dpi = 300)

# Health  per age: 
# Filter and create age groups
age_data <- main %>%
  filter(!is.na(age)) %>%  # Remove NA values in the age column
  mutate(
    age_group = cut(
      age,
      breaks = c(25, 30, 35, 40, 45, 50, 55),  # Define the breaks explicitly
      include.lowest = TRUE,  # Ensures 25 is included in the first group
      right = TRUE,  # Ensures 55 is included in the last group
      labels = c("[25-30)", "[30-35)", "[35-40)", "[40-45)", "[45-50)", "[50-55]")  # Labels with brackets
    )
  )

ggplot(age_data, aes(x = age_group, fill = age_group)) +
  geom_bar(alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette = "Blues", guide = "none") +  # Use a blue color palette
  labs(
    title = "Age Group Distribution",
    x = "Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
ggsave("age2.png", plot = last_plot(), dpi = 300)


# Create violin plot for health by 5-year age groups
ggplot(age_data, aes(x = age_group, y = health, fill = age_group)) +
  geom_violin(trim = FALSE, alpha = 0.7, color = "black") +  # Add transparency and black border
  scale_fill_brewer(palette = "Blues", guide = "none") +  # Use a blue color palette
  labs(
    title = "Distribution of Health by 5-Year Age Groups",
    x = "Age Group",
    y = "Health"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
ggsave("age_health.png", plot = last_plot(), dpi = 300)

# Graph for health_decline_2yrs per age: 
ggplot(age_data, aes(x = age_group, fill = factor(health_decline_2yrs))) +
  geom_bar(position = "fill", alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_manual(
    values = c("#A6CEE3", "#1E2B4F"),  # Light blue and dark blue
    labels = c("No Decline", "Decline"),
    name = "Health Decline"
  ) +
  labs(
    title = "Proportion of Health Decline by Age Group",
    subtitle = "Stacked Bar Chart Showing Health Decline Across Age Groups",
    x = "Age Group",
    y = "Proportion",
    fill = "Health Decline"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "top",  # Move legend to the top
    legend.text = element_text(size = 12, color = "#1E2B4F"),  # Dark blue legend text
    legend.title = element_text(size = 14, face = "bold", color = "#1E2B4F")  # Bold and dark blue legend title
  )
ggsave("age_health_decline.png", plot = last_plot(), dpi = 300)


# Graph for worr_health_dummy per age:
ggplot(age_data, aes(x = age_group, fill = factor(worr_health_dummy))) +
  geom_bar(position = "fill", alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_manual(
    values = c("#A6CEE3", "#1E2B4F"),  # Light blue and dark blue
    labels = c("No Worries", "Worries"),
    name = "Health Worry"
  ) +
  labs(
    title = "Proportion of Health Worries by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Health Worry"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "top",  # Move legend to the top
    legend.text = element_text(size = 12, color = "#1E2B4F"),  # Dark blue legend text
    legend.title = element_text(size = 14, face = "bold", color = "#1E2B4F")  # Bold and dark blue legend title
  )

ggsave("age_worr_health.png", plot = last_plot(), dpi = 300)

# Graph for health_satisfaction per age: 
ggplot(age_data, aes(x = age_group, y = health_satisfaction, fill = age_group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black border
  scale_fill_brewer(palette = "Blues", guide = "none") +  # Use a blue color palette
  labs(
    title = "Health Satisfaction by Age Group",
    x = "Age Group",
    y = "Health Satisfaction"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
ggsave("age_health_satisfaction.png", plot = last_plot(), dpi = 300)

```
In our dataset, the most participants we have are around the age of 45. For all health variables, we notice the same trend: the positive subjective health perceptions are higher within younger individuals, health worries increases with age, and health satisfaction decreases with age. We also notice that health decline is stable accross all age groups at around 24%. Because we expect older individuals to have a higher income, we should look into the intersection of age and income and their impact on health. 

# Scatter Plot of Net Income vs. Age
```{r}
ggplot(main, aes(x = age, y = net_income)) +
  geom_point(alpha = 0.6, color = "#1E2B4F") +  # Semi-transparent points with teal color
  geom_smooth(method = "loess", color = "#F00000", se = TRUE, size = 1.2) +  # Add a red LOESS trend line
  labs(
    title = "Net Income vs. Age",
    x = "Age",
    y = "Net Income"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space around the plot
  )

ggsave("income_age.png", plot = last_plot(), dpi = 300)
```
# Net income does increase with age but not significantly. 

# Marital Status Disparities

```{r}
# Group data by pid and rel_status, and summarize
main_rel <- main %>%
  group_by(pid, rel_status) %>%
  summarize(
    health = mean(health, na.rm = TRUE),
    health_decline_2yrs = mean(health_decline_2yrs, na.rm = TRUE),
    worr_health_dummy = mean(worr_health_dummy, na.rm = TRUE),
    health_satisfaction = mean(health_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )
# Distribution of rel_status variable
ggplot(main_rel, aes(x = factor(rel_status), fill = factor(rel_status))) +
  geom_bar(alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette = "RdBu", guide = "none") +  
  labs(
    title = "Distribution of Relationship Status",
    x = "Relationship Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

ggsave("rel_status.png", plot = last_plot(), dpi = 300)

# Health  per rel_status: 

ggplot(main_rel, aes(x = factor(rel_status), y = health, fill = factor(rel_status))) +
  geom_bar(stat = "summary", fun = "mean", alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette = "RdBu", guide = "none") +  
  labs(
    title = "Average Health by Relationship Status",
    x = "Relationship Status",
    y = "Average Health"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )


ggsave("rel_status_health.png", plot = last_plot(), dpi = 300)

# Graph for health_decline_2yrs per rel_status: 
ggplot(main_rel, aes(x = factor(rel_status), y = health_decline_2yrs, fill = factor(rel_status))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black border
  scale_fill_brewer(palette = "RdBu", guide = "none") +  
  labs(
    title = "Health Decline by Relationship Status",
    x = "Relationship Status",
    y = "Health Decline"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )


ggsave("rel_status_health_decline.png", plot = last_plot(), dpi = 300)


# Graph for worr_health_dummy per rel_status:
ggplot(main_rel, aes(x = factor(rel_status), y = worr_health_dummy, fill = factor(rel_status))) +
  geom_bar(stat = "summary", fun = "mean", alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette ="RdBu", guide = "none") + 
  labs(
    title = "Average Health Worries by Relationship Status",
    x = "Relationship Status",
    y = "Average Health Worries"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 50, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )


ggsave("rel_status_worr_health.png", plot = last_plot(), dpi = 300)

# Graph for health_satisfaction per rel_status: 
ggplot(main_rel, aes(x = factor(rel_status), y = health_satisfaction, fill = factor(rel_status))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette = "Set2", guide = "none") +  # Use the Set2 palette for soft colors
  labs(
    title = "Health Satisfaction by Relationship Status",
    x = "Relationship Status",
    y = "Health Satisfaction"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

ggsave("rel_status_health_satisfaction.png", plot = last_plot(), dpi = 300)
```
Our dataset contains mostly married individuals, which makes sense as our main population is around the age of 45. Average health is quite consistent across all relationship statuses, as well as health decline and health satisfaction. It is different for separated individuals and those with spouses abroad but the low number of these observations causing high variance does not allow us to make conclusions. 

We notice that single individuals are the least worried about their health but we expect this to be caused by the younger age of single participants. We will therefore examine the relationship between marital status and age.


# Box Plot of Age by Relationship Status
```{r}
ggplot(main, aes(x = factor(rel_status), y = age, fill = factor(rel_status))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +
  scale_x_discrete(
    labels = c(
      "1" = "Single",
      "2" = "Married",
      "3" = "Divorced",
      "4" = "Widowed"
    )
  ) +
  scale_fill_brewer(palette = "RdBu", guide = "none") +  # Use the RdBu palette
  labs(
    title = "Age by Relationship Status",
    x = "Relationship Status",
    y = "Age"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

ggsave("rel_status_age.png", plot = last_plot(), dpi = 300)

```

As expected, the single group is younger but this is not significant as the variance is quite high.

# Germborn Disparities

```{r}
# Group data by pid and germborn, and summarize
main_germborn <- main %>%
  group_by(pid, germborn) %>%
  summarize(
    health = mean(health, na.rm = TRUE),
    health_decline_2yrs = mean(health_decline_2yrs, na.rm = TRUE),
    worr_health_dummy = mean(worr_health_dummy, na.rm = TRUE),
    health_satisfaction = mean(health_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )
# Distribution of germborn variable
ggplot(main_germborn, aes(x = factor(germborn), fill = factor(germborn))) +
  geom_bar(alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_x_discrete(
    labels = c(
      "0" = "Born outside Germany", 
      "1" = "Born in Germany"
    )
  ) +
  scale_fill_manual(
    values = c("#A6CEE3", "#1E2B4F"), 
    guide = "none"  # Remove legend
  ) +
  labs(
    title = "Distribution of German-Born Individuals",
    x = "Migration Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space around the plot
  )
ggsave("germborn.png", plot = last_plot(), dpi = 300)


# Health per germborn: 
ggplot(main_germborn, aes(x = factor(germborn), y = health, fill = factor(germborn))) +
  geom_violin(trim = FALSE, alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_manual(
    values = c("#A6CEE3", "#1E2B4F"),  # Light blue and dark blue
    labels = c("Born Outside of Germany", "Born in Germany"),
    name = "German-Born Status"
  ) +
  scale_x_discrete(
    labels = c(
      "0" = "Born Outside of Germany",
      "1" = "Born in Germany"
    )
  ) +
  labs(
    title = "Distribution of Health by Migration Status",
    x = "",
    y = "Health"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
ggsave("germborn_health.png", plot = last_plot(), dpi = 300)

# Graph for health_decline_2yrs per germborn: 
ggplot(main_germborn, aes(x = factor(germborn), y = health_decline_2yrs, fill = factor(germborn))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black borders
  scale_fill_manual(
    values = c("#A6CEE3", "#1E2B4F"),  # Light blue and dark blue
    labels = c("Born Outside of Germany", "Born in Germany"),
    name = "Migration Status"
  ) +
  scale_x_discrete(
    labels = c(
      "0" = "Born Outside of Germany",
      "1" = "Born in Germany"
    )
  ) +
  labs(
    title = "Health Decline by Migration Status",
    x = "",
    y = "Health Decline"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F", hjust = 0.5),  # Centered and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )
ggsave("germborn_health_decline.png", plot = last_plot(), dpi = 300)

# Graph for worr_health_dummy per germborn:
ggplot(main_germborn, aes(x = worr_health_dummy, fill = factor(germborn))) +
  geom_density(alpha = 0.5, color = "black") +  # Add transparency and black outline for density curves
  scale_fill_manual(
    values = c("#F00000", "#1E2B4F"), 
    labels = c("Born Outside of Germany", "Born in Germany"),
    name = "Migration Status"
  ) +
  labs(
    title = "Density of Health Worries by Migration Status",
    x = "Health Worries",
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "top",  # Place legend at the top
    legend.text = element_text(size = 12, color = "#1E2B4F"),  # Dark blue legend text
    legend.title = element_text(size = 14, face = "bold", color = "#1E2B4F")  # Bold and dark blue legend title
  )

ggsave("germborn_worr_health.png", plot = last_plot(), dpi = 300)


# Graph for health_satisfaction per germborn: 
ggplot(main_germborn, aes(x = factor(germborn), y = health_satisfaction, fill = factor(germborn))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black borders
  scale_fill_manual(
    values = c("#A6CEE3", "#1E2B4F"),  # Light blue and dark blue
    labels = c("Born Outside of Germany", "Born in Germany"),
    name = "Migration Status"
  ) +
  scale_x_discrete(
    labels = c(
      "0" = "Born Outside of Germany",
      "1" = "Born in Germany"
    )
  ) +
  labs(
    title = "Health Satisfaction by Migration Status",
    x = "",
    y = "Health Satisfaction"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F", hjust = 0.5),  # Bold and centered dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

ggsave("germborn_health_satisfaction.png", plot = last_plot(), dpi = 300)

```
Our dataset mostly contains individuals born in Germany, which makes sense as it is the SOEP data. For people born in Germany or Outside of Germany, health distributions are quite similar. We do notice that individuals born outside of Germany have smaller health declines but also higher health worries and lower health satisfaction. The small differences, the contradictory findings, merged with the fact that we only have a small sample of individuals born outside of Germany does not allow us to make conclusion. 

# Relationship between subjective and objective health

```{r}
# worr_health_categorical per health_satisfaction: 
ggplot(main, aes(x = factor(worr_health_categorical), y = health_satisfaction, fill = factor(worr_health_categorical))) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette = "RdBu", guide = "none") +  # Use the Blues palette for a gradient
  scale_x_discrete(
    labels = c(
      "1" = "Does Not Worry",
      "2" = "Worries a Little",
      "3" = "Worries a Lot"
    )
  ) +
  labs(
    title = "Health Satisfaction by Health Worry Categories",
    x = "Health Worry Category",
    y = "Health Satisfaction"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )

ggsave("worr_health_health_satisfaction.png", plot = last_plot(), dpi = 300)


# health_satisfaction per life_satisfaction: 
ggplot(main, aes(x = life_satisfaction, y = health_satisfaction)) +
  geom_bin2d(bins = 30) +
  scale_fill_gradient(
    low = "#A6CEE3",  # Light blue
    high = "#1E2B4F", # Dark blue
    name = "Count"    # Legend title
  ) +
  labs(
    title = "Health Satisfaction vs. Life Satisfaction",
    x = "Life Satisfaction",
    y = "Health Satisfaction"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "right",  # Position legend on the right
    legend.text = element_text(size = 12, color = "#1E2B4F"),  # Dark blue legend text
    legend.title = element_text(size = 14, face = "bold", color = "#1E2B4F")  # Bold and dark blue legend title
  )

ggsave("life_health_satisfaction.png", plot = last_plot(), dpi = 300)

# health per BMI graph: 
ggplot(main, aes(x = factor(bmi_categorical), y = health, fill = factor(bmi_categorical))) +
  geom_violin(trim = FALSE, alpha = 0.7, color = "black") +  # Add transparency and black borders
  scale_fill_brewer(palette = "Blues", guide = "none") +  # Use the Blues palette
  scale_x_discrete(
    labels = c(
      "1" = "Underweight",
      "2" = "Normal Weight",
      "3" = "Overweight",
      "4" = "Obese"
    )
  ) +
  labs(
    title = "Health Distribution by BMI Category",
    x = "BMI Category",
    y = "Health"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, face = "bold", color = "#1E2B4F"),  # Tilted and bold dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Legend styling
    legend.position = "none"  # Remove legend
  )


ggsave("health_bmi.png", plot = last_plot(), dpi = 300)



```
Further analyses to confirm our expectations, we see that the higher the health satisfaction, the lower the individuals worry about their health and that goes both ways. 

Similarly, the higher life satisfaction, the higher health satisfaction as we observe a linear increasing relationship between the two. 

Analyzing BMI, we also see that normal and overweight individuals show higher health. However, due to the low amount of observations of underweight and obese individuals, we cannot make conclusions.

# Detailed Summary Statistics Tables

```{r}
# Define health group labels and titles
main <- main %>%
  mutate(
    health_group = case_when(
      health >= 1 & health <= 2 ~ "1-2",
      health > 2 & health <= 3 ~ "2-3",
      health > 3 & health <= 5 ~ "4-5",
      TRUE ~ NA_character_
    )
  )

health_titles <- c(
  "1-2" = "Poor health",
  "2-3" = "Moderate health",
  "4-5" = "Good health"
)

# Function to calculate summary statistics for a variable
calculate_group_summary <- function(data, var_name, var_label) {
  data %>%
    summarise(
      Mean = mean(!!sym(var_name), na.rm = TRUE),
      SD = sd(!!sym(var_name), na.rm = TRUE),
      Min = min(!!sym(var_name), na.rm = TRUE),
      Max = max(!!sym(var_name), na.rm = TRUE),
      Count = n()
    ) %>%
    mutate(Variable = var_label)
}

# Variables to calculate statistics for
variables <- list(
  list(name = "BMI", label = "BMI"),
  list(name = "worr_health_dummy", label = "Worries About Health (Dummy)"),
  list(name = "smoking_dummy", label = "Smoking (Dummy)"),
  list(name = "health_in_2yrs", label = "Health in 2 Years"),
  list(name = "health_decline_2yrs", label = "Expected Health Decline in 2 Years"),
  list(name = "life_satisfaction", label = "Life Satisfaction"),
  list(name = "health_satisfaction", label = "Health Satisfaction"),
  list(name = "tenure", label = "Tenure"),
  list(name = "net_income", label = "Net Income"),
  list(name = "work_time_weekly", label = "Work Time Weekly"),
  list(name = "total_years_unemployment", label = "Total Years Unemployment"),
  list(name = "worr_job_dummy", label = "Job Worries (Dummy)"),
  list(name = "worr_financial_dummy", label = "Financial Worries (Dummy)"),
  list(name = "worr_economic_dummy", label = "Economic Worries (Dummy)"),
  list(name = "male", label = "Gender (Male = 1)"),
  list(name = "age", label = "Age"),
  list(name = "educ", label = "Education Level")
  
)

# Initialize a list to store tables for each group
group_tables <- list()

# Generate separate tables for each health group
for (group in unique(main$health_group)) {
  group_data <- main %>% filter(health_group == group)
  group_table <- data.frame(
    Variable = character(),
    Mean = numeric(),
    SD = numeric(),
    Min = numeric(),
    Max = numeric(),
    Count = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (var in variables) {
    stats <- calculate_group_summary(group_data, var$name, var$label)
    if (!is.null(stats)) {
      group_table <- bind_rows(group_table, stats)
    }
  }
  
  # Round all numeric columns to 1 digit after the decimal point
  group_table <- group_table %>%
    mutate(
      Mean = round(Mean, 1),
      SD = round(SD, 1),
      Min = round(Min, 1),
      Max = round(Max, 1)
    )
  
  # Add the table to the list with the title as the key
  table_title <- health_titles[group]
  group_tables[[table_title]] <- group_table
}

# Display each group's table with its title
for (title in names(group_tables)) {
  print(paste("Summary statistics for:", title))
  print(group_tables[[title]])
}

# Save each group's table to separate CSV files
for (title in names(group_tables)) {
  write.csv(group_tables[[title]], paste0("summary_table_", gsub(" ", "_", title), ".csv"), row.names = FALSE)
}

```


# Short Summary Statistics Table

```{r}
# Define health group labels and titles
main <- main %>%
  mutate(
    health_group = case_when(
      health >= 1 & health <= 2 ~ "1-2",
      health > 2 & health <= 3 ~ "2-3",
      health > 3 & health <= 5 ~ "4-5",
      TRUE ~ NA_character_
    )
  )

health_titles <- c(
  "1-2" = "Poor health",
  "2-3" = "Moderate health",
  "4-5" = "Good health"
)

# Variables to calculate statistics for
variables <- list(
list(name = "BMI", label = "BMI"),
  list(name = "worr_health_dummy", label = "Worries About Health (Dummy)"),
  list(name = "smoking_dummy", label = "Smoking (Dummy)"),
  list(name = "health_in_2yrs", label = "Health in 2 Years"),
  list(name = "health_decline_2yrs", label = "Expected Health Decline in 2 Years"),
  list(name = "life_satisfaction", label = "Life Satisfaction"),
  list(name = "health_satisfaction", label = "Health Satisfaction"),
  list(name = "tenure", label = "Tenure"),
  list(name = "net_income", label = "Net Income"),
  list(name = "work_time_weekly", label = "Work Time Weekly"),
  list(name = "total_years_unemployment", label = "Total Years Unemployment"),
  list(name = "employment", label = "Employment Status (Employed = 1)"),
  list(name = "worr_job_dummy", label = "Job Worries (Dummy)"),
  list(name = "worr_financial_dummy", label = "Financial Worries (Dummy)"),
  list(name = "worr_economic_dummy", label = "Economic Worries (Dummy)"),
  list(name = "male", label = "Gender (Male = 1)"),
  list(name = "age", label = "Age"),
  list(name = "educ", label = "Education Level"),
  list(name = "university", label = "University Degree (Dummy)")
)

# Create a summary table
summary_table <- data.frame(Variable = sapply(variables, function(x) x$label))

# Calculate means for each health category
for (group in unique(main$health_group)) {
  group_data <- main %>% filter(health_group == group)
  group_means <- sapply(variables, function(var) mean(group_data[[var$name]], na.rm = TRUE))
  summary_table[[health_titles[group]]] <- round(group_means, 1)
}

# Add row for the number of observations in each health category
num_observations <- sapply(unique(main$health_group), function(group) nrow(main %>% filter(health_group == group)))
summary_table <- rbind(summary_table, c("Number of Observations", num_observations))

# Display the summary table
print(summary_table)

# Save the summary table as a CSV file
write.csv(summary_table, "summary_table_short.csv", row.names = FALSE)

```

The table provides descriptive statistics for individuals categorized by their self-reported health status: Good health, Moderate health, and Poor health. Each variable is summarized for these three health categories, allowing us to analyze how different characteristics vary across health statuses. 
- BMI	Individuals with poorer health tend to have higher BMI (Good: 25.3 → Poor: 27.8), suggesting a link between obesity and poorer health.
- Worries About Health (Dummy)	Health worries increase with poorer health (Good: 0.5 → Poor: 1.0), as expected.
- Individuals in poorer health have significantly lower expectations of good health in the future (Good: 3.9 → Poor: 2.6), also as expected.
- Similarly, individuals in Poor health report much lower health satisfaction (Good: 8.1 → Poor: 3.8) and a decline in life satisfaction as health worsens (Good: 7.8 → Poor: 5.7).
- Higher unemployment duration is associated with poorer health (Good: 0.7 years → Poor: 2.1 years).
- Worries about finances and the economy increase slightly as health worsens.
- Individuals in Good health have significantly higher incomes (Good: €1574 → Poor: €1060).
- Poor health correlates with increasing age (Good: 39.8 → Poor: 43.3).
- Education level decreases with poorer health (Good: 4.2 → Poor: 3.6). Lower education might contribute to worse health outcomes.
- Surprisingly, those in poorer health report less expected decline (Good: 0.3 → Poor: 0.1). This could indicate resignation or lower health awareness.

But, in general, in our dataset, ,ost individuals report Good health (45,372), with fewer in Moderate health (21,859) and the least in Poor health (9,276). This indicates a skewed distribution towards better health.


# Modelling

##  Lasso Regression 

```{r}
# Convert data to panel format
panel_data <- pdata.frame(main, index = c("pid", "syear"))  # pid = individual ID, syear = time

# Prepare data for LASSO regression
lasso_data <- panel_data %>%
  dplyr::select(
    pid, syear, health_decline_2yrs, health_in_2yrs, worr_health_dummy, health, BMI, smoking_dummy, life_satisfaction, health_satisfaction,
    tenure, net_income, work_time_weekly, total_years_unemployment, 
    worr_job_dummy, worr_financial_dummy, worr_economic_dummy,
    male, age, educ
  ) %>%
  na.omit()  # Remove missing values

# Check class distribution in y
class_distribution <- table(lasso_data$health_decline_2yrs)
print(class_distribution)

# Split the data into training and test data
set.seed(123) # For reproducibility
trainIndex <- createDataPartition (lasso_data$health_in_2yrs, p = 0.7, list = FALSE)
# Convert trainIndex to a vector
trainIndex <- as.vector(trainIndex)
# Use the vector to split the data
trainData <- lasso_data[trainIndex, ]
testData <- lasso_data[-trainIndex, ]
dim(trainData)

#Standardize variables 
# Define relevant numerical variables as a character vector
numerical_vars <- c(
  "BMI", "smoking_dummy", "life_satisfaction", "health_satisfaction",
  "tenure", "net_income", "work_time_weekly", "total_years_unemployment",
  "worr_job_dummy", "worr_financial_dummy", "worr_economic_dummy",
  "male", "age", "educ", "worr_health_dummy", "health"
)

# Apply pre-processing (center and scale) only to the selected numerical columns
preProcValues <- preProcess(trainData[, numerical_vars], method = c("center", "scale"))

# Standardize the training data 
trainData[, numerical_vars] <- predict(preProcValues, trainData[, numerical_vars])

# Standardize the test data
testData[, numerical_vars] <- predict(preProcValues, testData[, numerical_vars])

# Calculate the number of points in train and test subsets
train_size <- nrow(trainData)
test_size <- nrow(testData)

# Create a data frame with the counts for waffle plot
data_proportion <- c(`Train Data` = train_size, `Test Data` = test_size)

# Create the waffle plot
waffle(data_proportion / sum(data_proportion) * 100, 
       rows = 10, # Adjust number of rows for granularity
       title = "Proportion of Data Points in Training and Test Sets",
       colors = c("#1E2B4F", "#F00000"))

set.seed(12345)

# Set up the training control
# Exclude pid and syear to avoid overfitting
trainData <- trainData %>% select(-pid, -syear, -health_decline_2yrs)
testData <- testData %>% select(-pid, -syear, -health_decline_2yrs)

# Define cross-validation as the method and 10 folds 
train_control <- trainControl(method = "cv", number = 10)

# Train the LASSO model
lasso_model <- train(
  # use 'shealth_decline_2yrs' as the dependent and all others as independent variables
  health_in_2yrs ~ ., 
  # specify the traininf data
  data = trainData,
  # We want to use the LASSO method (glmnet)
  method = "glmnet",
  # Use the predefined training specification
  trControl = train_control,
  # Define what values for lambda we should try (tuning)
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0.001, 0.5, by = 0.001))  
)

# Display the best lambda value and model summary
print(lasso_model$bestTune)
# Visualize the cross-validation results
ggplot(lasso_model) +
  labs(
    title = "LASSO Model Cross-Validation Results",
    subtitle = "RMSE vs. Log(Lambda) Across Cross-Validation Folds",
    x = "Log(Lambda)",  # X-axis represents lambda values (log-scaled)
    y = "RMSE"          # Y-axis represents cross-validated RMSE
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis labels
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis labels
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 20, b = 20, l = 20)  # Add spacing around the plot
  )
#Save Graph
ggsave("lasso_plot.png", plot = last_plot(), width = 16, height = 8, dpi = 300)
# Model Evaluation
# Predict on the test set
predictions <- predict(lasso_model, newdata = testData)

# Calculate RMSE and R-squared
rmse <- RMSE(predictions, testData$health_in_2yrs)
r2 <- R2(predictions, testData$health_in_2yrs)

# Print the evaluation metrics
print(paste0("RMSE: ", rmse))
print(paste0("R-squared: ", r2))

# Extract the non-zero coefficients of the best model
coefficients_regression <- coef(lasso_model$finalModel, s = lasso_model$bestTune$lambda)
non_zero_coefficients_regression <- coefficients_regression[coefficients_regression != 0]
print(non_zero_coefficients_regression)
```

```{r}
#Plotting coefficients
# Convert sparse matrix to data frame and filter non-zero coefficients
non_zero_coefficients <- data.frame(
  Variable = rownames(as.matrix(coefficients_regression)),
  Coefficient = as.numeric(as.matrix(coefficients_regression))
) %>%
  filter(Coefficient != 0)  # Keep only non-zero coefficients

# Remove the intercept and rename variables
non_zero_coefficients <- non_zero_coefficients %>%
  filter(Variable != "(Intercept)") %>%  # Remove the intercept
  mutate(
    Variable = case_when(  # Rename variables for cleaner labels
      Variable == "health" ~ "Health",
      Variable == "health_satisfaction" ~ "Health Satisfaction",
      Variable == "life_satisfaction" ~ "Life Satisfaction",
      Variable == "educ" ~ "Education",
      Variable == "male" ~ "Gender (Male)",
      Variable == "worr_job_dummy" ~ "Job Worries",
      Variable == "tenure" ~ "Tenure",
      Variable == "work_time_weekly" ~ "Weekly Work Hours",
      Variable == "net_income" ~ "Net Income",
      Variable == "total_years_unemployment" ~ "Total Years Unemployed",
      Variable == "worr_financial_dummy" ~ "Financial Worries",
      Variable == "smoking_dummy" ~ "Smoking",
      Variable == "worr_health_dummy" ~ "Health Worries",
      Variable == "BMI" ~ "BMI",
      Variable == "age" ~ "Age",
      TRUE ~ Variable  # Keep any variable not explicitly renamed
    )
  )

# Plotting non-zero coefficients with cleaner labels
ggplot(non_zero_coefficients, aes(x = reorder(Variable, Coefficient), y = Coefficient, fill = Coefficient > 0)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Non-Zero Coefficients from Lasso Regression",
    x = "",
    y = "Coefficient Value"
  ) +
  scale_fill_manual(
    values = c("#FF0000", "#163E64"),  # Red for negative, dark blue for positive
    guide = "none"  # Remove the legend
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),  # Remove y-axis title
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis text
    axis.text.y = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis text

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space on the right and top/bottom
  )


#Save Graph
ggsave("lasso_coefficients.png", plot = last_plot(), width = 16, height = 8, dpi = 300)

# Visualize the lasso model
plot(lasso_model$finalModel, xvar = "lambda", label = TRUE)


```

## Interpretation


## Lasso Classification

```{r}
# Split the data into training and test data
set.seed(123) # For reproducibility
trainIndex <- createDataPartition(lasso_data$health_decline_2yrs, p = 0.7, list = FALSE)
# Convert trainIndex to a vector
trainIndex <- as.vector(trainIndex)
# Use the vector to split the data
trainData <- lasso_data[trainIndex, ]
testData <- lasso_data[-trainIndex, ]

# Standardize numerical variables
# Define relevant numerical variables as a character vector
numerical_vars <- c(
  "BMI", "smoking_dummy", "life_satisfaction", "health_satisfaction",
  "tenure", "net_income", "work_time_weekly", "total_years_unemployment",
  "worr_job_dummy", "worr_financial_dummy", "worr_economic_dummy",
  "male", "age", "educ", "worr_health_dummy", "health"
)

# Apply pre-processing (center and scale) only to the selected numerical columns
preProcValues <- preProcess(trainData[, numerical_vars], method = c("center", "scale"))

# Standardize the training data
trainData[, numerical_vars] <- predict(preProcValues, trainData[, numerical_vars])

# Standardize the test data
testData[, numerical_vars] <- predict(preProcValues, testData[, numerical_vars])

# Exclude pid and syear to avoid overfitting
trainData <- trainData %>% select(-pid, -syear, -health_in_2yrs)
testData <- testData %>% select(-pid, -syear, health_in_2yrs)

# Convert health_decline_2yrs to a factor
trainData$health_decline_2yrs <- factor(trainData$health_decline_2yrs, levels = c(0, 1), labels = c("No", "Yes"))
testData$health_decline_2yrs <- factor(testData$health_decline_2yrs, levels = c(0, 1), labels = c("No", "Yes"))

# Set up the training control
train_control <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

# Train the LASSO model for classification
set.seed(12345)
lasso_model <- train(
  health_decline_2yrs ~ .,  # Dependent variable
  data = trainData,
  method = "glmnet",        # Specify LASSO
  family = "binomial",      # Classification for binary outcome
  trControl = train_control,  # Use predefined training control
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0.001, 0.5, by = 0.001)), # LASSO tuning grid
  metric = "ROC"  # Use ROC as the evaluation metric
)

# Display the best lambda value and model summary
print(lasso_model$bestTune)
# Visualize the cross-validation results
plot(lasso_model)

# Extract the cross-validation results
cv_results <- lasso_model$results

# Create a ggplot visualization
ggplot(cv_results, aes(x = log(lambda), y = ROC, color = ROC)) +
  geom_point(size = 2, alpha = 0.8) +  # Add points for each lambda value
  geom_line(size = 1) +  # Add a line to connect points
  labs(
    title = "LASSO Model Cross-Validation Results",
    subtitle = "ROC vs. Log(Lambda) Across Cross-Validation Folds",
    x = "Log(Lambda)",
    y = "ROC"
  ) +
  scale_color_gradient(low = "#A6CEE3", high = "#1E2B4F") +  # Gradient from light blue to dark blue
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis text
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis text
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title and subtitle styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold subtitle
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20),  # Add more space around the plot
    
    # Remove legend
    legend.position = "none" 

  )

#Save Graph
ggsave("lasso_model_classification.png", plot = last_plot(), width = 16, height = 8, dpi = 300)

# Model Evaluation
# Predict probabilities on the test set
predictions <- predict(lasso_model, newdata = testData, type = "prob")

# Predict binary outcomes (threshold = 0.5)
predicted_classes <- ifelse(predictions[, "Yes"] > 0.5, "Yes", "No")

# Confusion matrix
conf_matrix <- confusionMatrix(factor(predicted_classes, levels = c("No", "Yes")), 
                               factor(testData$health_decline_2yrs, levels = c("No", "Yes")))
print(conf_matrix)

# Evaluate model performance
print(paste0("Accuracy: ", conf_matrix$overall["Accuracy"]))
print(paste0("Sensitivity: ", conf_matrix$byClass["Sensitivity"]))
print(paste0("Specificity: ", conf_matrix$byClass["Specificity"]))
```

```{r}
# Convert sparse matrix to data frame and filter non-zero coefficients
coefficients <- coef(lasso_model$finalModel, s = lasso_model$bestTune$lambda)

non_zero_coefficients <- data.frame(
  Variable = rownames(as.matrix(coefficients)),
  Coefficient = as.numeric(as.matrix(coefficients))
) %>%
  filter(Coefficient != 0)  # Keep only non-zero coefficients

# Remove the intercept and rename variables
non_zero_coefficients <- non_zero_coefficients %>%
  filter(Variable != "(Intercept)") %>%  # Remove the intercept
  mutate(
    Variable = case_when(  # Rename variables for cleaner labels
      Variable == "health" ~ "Health",
      Variable == "health_satisfaction" ~ "Health Satisfaction",
      Variable == "life_satisfaction" ~ "Life Satisfaction",
      Variable == "educ" ~ "Education Level",
      Variable == "male" ~ "Gender (Male=1)",
      Variable == "worr_job_dummy" ~ "Job Worries (Dummy)",
      Variable == "tenure" ~ "Tenure",
      Variable == "work_time_weekly" ~ "Work Time Weekly",
      Variable == "net_income" ~ "Net Income",
      Variable == "total_years_unemployment" ~ "Total Years Unemployed",
      Variable == "worr_financial_dummy" ~ "Financial Worries (Dummy)",
      Variable == "smoking_dummy" ~ "Smoking (Dummy)",
      Variable == "worr_health_dummy" ~ "Worries About Health (Dummy)",
      Variable == "BMI" ~ "BMI",
      Variable == "age" ~ "Age",
      TRUE ~ Variable  # Keep any variable not explicitly renamed
    )
  )

# Plotting non-zero coefficients with cleaner labels
ggplot(non_zero_coefficients, aes(x = reorder(Variable, Coefficient), y = Coefficient, fill = Coefficient > 0)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Non-Zero Coefficients from Lasso Classification",
    x = "",
    y = "Coefficient Value"
  ) +
  scale_fill_manual(
    values = c("#FF0000", "#163E64"),  # Red for negative, dark blue for positive
    guide = "none"  # Remove the legend
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.title.y = element_blank(),  # Remove y-axis title
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis text
    axis.text.y = element_text(size = 12, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis text

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space around the plot
  )
#Save Graph
ggsave("lasso_coefficients_classification.png", plot = last_plot(), width = 16, height = 8, dpi = 300)

```

## Random Forest Classification

```{r}
# Create time-lagged variables
panel_data <- panel_data %>%
  group_by(pid) %>%
  arrange(syear) %>%
  mutate(
    health_lag1 = lag(health, 1),  # Previous year's health
    net_income_lag1 = lag(net_income, 1),  # Previous income
    work_time_weekly_lag1 = lag(work_time_weekly, 1),
    tenure_lag1 = lag(tenure, 1)
  ) %>%
  ungroup()

# Define relevant numerical variables as a character vector
numerical_vars <- c(
  "BMI", "smoking_dummy", "life_satisfaction", "health_satisfaction",
  "tenure", "net_income", "work_time_weekly", "total_years_unemployment",
  "worr_job_dummy", "worr_financial_dummy", "worr_economic_dummy",
  "male", "age", "educ", "worr_health_dummy", "health","health_lag1","net_income_lag1",
  "work_time_weekly_lag1","tenure_lag1"
)

# Create rf_data by removing specific columns
rf_data <- panel_data %>%
  dplyr::select(-pid, -syear, -health_in_2yrs)

# Ensure health_decline_2yrs Is a Factor
rf_data$health_decline_2yrs <- factor(rf_data$health_decline_2yrs, levels = c(0, 1), labels = c("No", "Yes"))

# Dynamically create the formula
class_tree_formula <- as.formula(paste("health_decline_2yrs ~", paste(numerical_vars, collapse = " + ")))

# Fit the decision tree
class_tree <- rpart(
  formula = class_tree_formula,
  data = rf_data,
  method = "class",
  control = rpart.control(
    cp = 0.001,       # Lower complexity parameter to allow more splits
    minsplit = 10,    # Minimum number of observations required to split
    minbucket = 5,    # Minimum observations in a terminal node
    maxdepth = 10     # Maximum depth of the tree
  )
)

# Plot the decision tree
rpart.plot(class_tree, main = "Decision Tree: Classification")

# Plot the decision tree with explicit splits
rpart.plot(class_tree, main = "Decision Tree: Classification", type = 5)

# Building the classification model

set.seed(123) # For reproducibility
# Create a training and test dataset
trainIndex <- createDataPartition (rf_data$health_decline_2yrs, p = 0.7, list = FALSE)
trainData <- rf_data[trainIndex, ]
testData <- rf_data[-trainIndex, ]
dim(trainData)
dim(testData)

# Creates a data frame (tuning_grid) where trees ranges from 5 to 400 in 
# steps of 5. The rmse column is initialized with NA to be filled later.
tuning_grid <- expand.grid(
  trees = seq(5, 400, by = 5),
  rmse  = NA
)


for(i in seq_len(nrow(tuning_grid))) { # Iterates over each row of the tuning_grid.
  
  # Fits a random forest model 
  fit <- ranger(
  formula    = class_tree_formula, 
  data       = trainData, 
  num.trees  = tuning_grid$trees[i],
  mtry       = sqrt(5), 
  verbose    = FALSE,
  seed       = 123
  )
  # Stores the RMSE of the model 
  tuning_grid$rmse[i] <- sqrt(fit$prediction.error)
}

```

```{r}
# Plot the rmse against then number of trees
ggplot(tuning_grid, aes(trees, rmse)) + 
  geom_line()

# Plot the RMSE against the number of trees with enhanced theme
ggplot(tuning_grid, aes(x = trees, y = rmse)) + 
  geom_line(color = "#163E64", size = 1) +  # Dark blue line
  geom_point(color = "#1E2B4F", size = 2, alpha = 0.8) +  # Points for emphasis
  labs(
    title = "Random Forest Model Tuning: RMSE vs. Number of Trees",
    x = "Number of Trees",
    y = "RMSE"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    # Backgrounds
    panel.background = element_rect(fill = "white", color = NA),  # White background
    plot.background = element_rect(fill = "white", color = NA),  # White outer background

    # Gridlines
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines

    # Axis styling
    axis.text.x = element_text(size = 12, color = "#1E2B4F"),  # Dark blue x-axis text
    axis.text.y = element_text(size = 12, color = "#1E2B4F"),  # Dark blue y-axis text
    axis.title.x = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue x-axis title
    axis.title.y = element_text(size = 14, face = "bold", color = "#1E2B4F"),  # Bold and dark blue y-axis title

    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#1E2B4F", margin = ggplot2::margin(b = 10)),  # Centered and bold title
    
    # Adjust overall plot margins
    plot.margin = ggplot2::margin(t = 20, r = 100, b = 20, l = 20)  # Add more space around the plot
  )
#Save Graph
ggsave("rf_rmse.png", plot = last_plot(), width = 16, height = 8, dpi = 300)

```

```{r}
# Defines the training scheme to be 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

tune_grid <- expand.grid(
  mtry = seq(2,4,0.05),  # Number of variables randomly sampled at each split
  splitrule = "gini", # Number of trees in the forest
  min.node.size = c(1, 5, 10) # Minimum size of terminal nodes
)

# Train the Random Forest model using ranger in caret
set.seed(123) # Set seed for reproducibility 
rf_model <- train(
  class_tree_formula, 
  data = trainData, 
  method = "ranger",             
  trControl = train_control, 
  tuneGrid = tune_grid, 
  num.trees = 100 # that we got previously               
)

# Print the best model and results
print(rf_model)

# Make predictions on the test set
predictions <- predict(rf_model, newdata = testData)

```

```{r}
# Confusion matrix to evaluate performance
confusionMatrix(predictions, testData$health_decline_2yrs, positive = "Yes")


# Fit the final model 'rf_model'
rf_final <- ranger(
  formula    = class_tree_formula, 
  data       = trainData, 
  num.trees  = 100,           # Number of trees
  probability = TRUE,         # Predict probabilities
  importance = "impurity",    # Variable importance measure
  mtry       = 3.15,          # Number of variables randomly sampled at each split
  min.node.size = 10,         # Minimum node size
  seed       = 123            # For reproducibility
)

# Get feature importance
v1 <- vip(rf_final) # Calculate and plot the variable importance
v1$data # Print the variable importance values
# Plot of feature importance
v1_table <- v1$data

# Clean and rename variables
v1_table <- v1_table %>%
  filter(Variable != "(Intercept)") %>%  # Remove the intercept
  mutate(
    Variable = case_when(  # Rename variables for cleaner labels
      Variable == "health" ~ "Health",
      Variable == "health_lag1" ~ "Health of previous year",
      Variable == "health_satisfaction" ~ "Health Satisfaction",
      Variable == "life_satisfaction" ~ "Life Satisfaction",
      Variable == "educ" ~ "Education Level",
      Variable == "male" ~ "Gender (Male=1)",
      Variable == "worr_job_dummy" ~ "Job Worries (Dummy)",
      Variable == "tenure" ~ "Tenure",
      Variable == "tenure_lag1" ~ "Tenure of previous year",
      Variable == "work_time_weekly" ~ "Weekly Work Hours",
      Variable == "work_time_weekly_lag1" ~ "Weekly Work Hours of previous year",
      Variable == "net_income" ~ "Net Income",
      Variable == "net_income_lag1" ~ "Net Income of previous year",
      Variable == "total_years_unemployment" ~ "Total Years Unemployed",
      Variable == "worr_financial_dummy" ~ "Financial Worries (Dummy)",
      Variable == "smoking_dummy" ~ "Smoking",
      Variable == "worr_health_dummy" ~ "Health Worries (Dummy)",
      Variable == "BMI" ~ "BMI",
      Variable == "age" ~ "Age",
      TRUE ~ Variable  # Keep any variable not explicitly renamed
    )
  )

# Create the bar plot
ggplot(v1_table, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", position = "dodge", fill = "#1E2B4F", color = "black") +
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Variable Importance from Random Forest",
    x = "",
    y = "Importance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = NA), # Dark blue background
    plot.background = element_rect(fill = "white", color = NA), # Dark blue outer background
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(size = 12, color = "#1E2B4F"), 
    axis.text.y = element_text(size = 12, face = "bold", color = "#1E2B4F"), 
    plot.title = element_text(face = "bold", size = 16, hjust = 0.25, color = "#1E2B4F"), # Shift title left and color 
    plot.caption = element_text(color = "#1E2B4F"), 
    axis.title = element_text(color = "#1E2B4F"), 
    legend.position = "none"  # Remove the legend
  )
#Save Graph
ggsave("rf_importance.png", plot = last_plot(), width = 16, height = 8, dpi = 300)

```
## Partial Dependence

```{r}
# Calculate the Partial Dependence for health_lag1
pdp_health_lag1 <- partial(
  object = rf_final, 
  pred.var = "health_lag1", 
  train = trainData,
  prob = TRUE  
)

# Calculate the Partial Dependence for male
pdp_male <- partial(
  object = rf_final, 
  pred.var = "male", 
  train = trainData,
  prob = TRUE
)

# Calculate the Partial Dependence for bmi
pdp_bmi <- partial(
  object = rf_final, 
  pred.var = "BMI", 
  train = trainData,
  prob = TRUE
)

#Transform tenure_lag1 into numeric
trainData$tenure_lag1 <- as.numeric(trainData$tenure_lag1)

# Calculate the Partial Dependence for tenure
pdp_tenure <- partial(
  object = rf_final, 
  pred.var = "tenure_lag1", 
  train = trainData,
  prob = TRUE
)
#Transform life_satisfaction into numeric
trainData$life_satisfaction <- as.numeric(trainData$life_satisfaction)
# Calculate the Partial Dependence for Life Satisfaction
pdp_life <- partial(
  object = rf_final, 
  pred.var = "life_satisfaction", 
  train = trainData,
  prob = TRUE
)

# Calculate the Partial Dependence for age
pdp_age <- partial(
  object = rf_final, 
  pred.var = "age", 
  train = trainData,
  prob = TRUE
)
```

```{r}
# Plot the PDP for health_lag1
p1 <- autoplot(pdp_health_lag1) +
  ylab("Health Decline Prob")+
  xlab("Health in previous year")+
  theme_minimal()

# Plot the PDP for male
p2 <- autoplot(pdp_male) +
  ylab("Health Decline Prob")+
  xlab("Male")+
  theme_minimal()

# Plot the PDP for bmi
p3 <- autoplot(pdp_bmi) +
  ylab("Health Decline Prob")+
  xlab("BMI")+
  theme_minimal()

# Plot the PDP for tenure
p4 <- autoplot(pdp_tenure) +
  ylab("Health Decline Prob")+
  xlab("Tenure of previous year")+
  theme_minimal()

# Plot the PDP for life_satisfaction
p5 <- autoplot(pdp_life) +
  ylab("Health Decline Prob")+
  xlab("Life Satisfaction")+
  theme_minimal()

# Plot the PDP for age
p6 <- autoplot(pdp_age) +
  ylab("Health Decline Prob")+
  xlab("Age")+
  theme_minimal()

# Combine the plots
(p1 | p2 ) /
(p3 | p4 ) /
(p5 | p6 ) 

```
